---
title: "Control Flow in<br/>R & Python"
subtitle: "Lecture 03"
author: "Dr. Colin Rundel"
footer: "Sta 523 - Fall 2026"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
  warning: true
engine: knitr
---


```{r setup}
#| message: false
#| warning: false
#| include: false
options(
  width=80
)
```

```{python py_setup}
#| echo: false
import math
```


# Logical &<br/>comparison operators

## Comparison operators

The syntax is nearly identical. The key difference is that R's comparison operators are *vectorized* (element-wise, returning a logical vector), whereas Python compares two objects and returns a single `bool`.

<br/>

| Comparison               | R           | Python     |
|:-------------------------|:------------|:-----------|
| less than                | `x < y`     | `x < y`    |
| greater than             | `x > y`     | `x > y`    |
| less than or equal to    | `x <= y`    | `x <= y`   |
| greater than or equal to | `x >= y`    | `x >= y`   |
| equal to                 | `x == y`    | `x == y`   |
| not equal to             | `x != y`    | `x != y`   |
| membership               | `x %in% y`  | `x in y`   |


## Comparisons

R compares element by element, while Python compares the objects as a whole and uses `in` for membership.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = c("A", "B", "C")
x == "A"
x %in% c("A", "C")
c(1, 2, 3) == c(1, 2, 3)
c(1, 2, 3) < c(1, 3, 0)
```
:::

::: {.column width='50%'}
```{python}
x = ["A", "B", "C"]
x == "A"
"A" in x
[1, 2, 3] == [1, 2, 3]
[1, 2, 3] < [1, 3, 0]
```
:::
::::

::: {.aside}
Python's ordered comparisons of lists are *lexicographic* - more on this in a moment, and on recycling and vectorization next time.
:::


## Comparing strings

Both languages can order strings, but differently - Python compares by Unicode code point (so ASCII `A`-`Z` sort before ASCII `a`-`z`), while R uses the collation rules of the current locale (roughly alphabetical, with case as a tie breaker).

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
"A" < "B"
"Good" < "Goodbye"
"A" < "a"
"a" < "B"
"Z" < "a"
```
:::

::: {.column width='50%'}
```{python}
"A" < "B"
"Good" < "Goodbye"
"A" < "a"
"a" < "B"
"Z" < "a"
```
:::
::::

::: {.aside}
See `Sys.getlocale("LC_COLLATE")` and `?Comparison` for R's locale (in the `C` locale R also orders by code point), and `locale.strcoll()` / `locale.strxfrm()` for locale-aware ordering in Python. The same rules apply to `sort()` and `sorted()`.
:::


## Lexicographic ordering

Ordered comparisons of strings, and of lists and tuples in Python, are *lexicographic* (dictionary order) - elements are compared pairwise from the front and the first difference decides the result. If one sequence runs out first, it sorts before the longer one.

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
"card" < "cart"
"cat" < "card"
"car" < "card"
```
:::

::: {.column width='50%'}
```{python}
"card" < "cart"
"cat" < "card"
"car" < "card"
```
```{python}
[1, 2, 3] < [1, 3, 0]
(1, 2) < (1, 2, 5)
```
:::
::::

. . .

R applies this only to strings (using the character ordering from the previous slide) - comparing vectors, as we saw, is element-wise, not lexicographic.

::: {.aside}
Only the elements up to the first difference are ever compared - `[0, "a"] < [1, 2]` is `True`, but `[1, "a"] < [1, 2]` raises a `TypeError` since `"a" < 2` is undefined.
:::


## Logical operators

<br/>

| Operation    | R (vectorized)           | R (scalar)                    | Python    |
|:-------------|:-------------------------|:------------------------------|:----------|
| and          | `x & y`                  | `x && y`                      | `x and y` |
| or           | <code>x &#124; y</code>  | <code>x &#124;&#124; y</code> | `x or y`  |
| not          | `!x`                     |                               | `not x`   |
| exclusive or | `xor(x, y)`              |                               | `x != y`  |


::: {.aside}
Python's `&`, <code>&#124;</code>, and `~` are *bitwise* operators for integers; confusingly, `numpy` and `pandas` overload them as element-wise logical operators (more on this later). The `x != y` shorthand for exclusive-or assumes `x` and `y` are booleans; for arbitrary values use `bool(x) != bool(y)`.
:::


## Vectorized vs scalar

In R you choose between the vectorized and scalar forms; in base Python `and` / `or` are always scalar, though with lists they can misleadingly look vectorized.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c(TRUE, FALSE, TRUE)
y = c(FALSE, TRUE, TRUE)
```
```{r}
x | y
x & y
```
```{r}
#| error: true
x || y
```
:::

::: {.column width='50%'}
```{python}
x = [True, False, True]
y = [False, True, True]
```

```{python}
x or y  # returns x
x and y # returns y
```
:::
::::

. . .

::: {.aside}
Since R 4.3, `&&` and <code>&#124;&#124;</code> throw an error if either side has length greater than 1 (older versions silently used only the first element).
:::


## Short-circuit evaluation

`&&` / <code>&#124;&#124;</code> in R and `and` / `or` in Python evaluate their left operand first and only evaluate the right operand if the left one does not already decide the outcome.

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
FALSE && stop("Error")
TRUE || stop("Error")
```
```{r}
#| error: true
TRUE && stop("Error")
```
```{r}
#| error: true
FALSE || stop("Error")
```
:::

::: {.column width='50%'}
```{python}
False and 1/0
True or 1/0
```
```{python}
#| error: true
True and 1/0
```
```{python}
#| error: true
False or 1/0
```
:::
::::

## Eager evaluation

Unlike the short-circuiting operators, `&` / <code>&#124;</code> in both languages always evaluate both operands:

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
#| error: true
FALSE & stop("Error")
```
```{r}
#| error: true
TRUE | stop("Error")
```
:::

::: {.column width='50%'}
```{python}
#| error: true
False & 1/0
```
```{python}
#| error: true
True | 1/0
```
:::
::::


## Guarding with short-circuits

This makes them useful as *guards*, where an earlier condition rules out inputs that would cause a later condition to error.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
#| error: true
x = "abc"
abs(x) > 1
```
```{r}
is.numeric(x) && abs(x) > 1
```
:::

::: {.column width='50%'}
```{python}
#| error: true
x = "abc"
abs(x) > 1
```
```{python}
isinstance(x, (int, float)) and abs(x) > 1
```
:::
::::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
#| error: true
x = NULL
abs(x) > 1
```
```{r}
!is.null(x) && abs(x) > 1
```
:::

::: {.column width='50%'}
```{python}
#| error: true
x = None
abs(x) > 1
```
```{python}
x is not None and abs(x) > 1
```
:::
::::


## Truthiness and short-circuiting

::: {.medium}
Python's `and` and `or` accept any values, not just `bool`s. They check the *truthiness* of `x` and short-circuiting decides which value comes back:

* `x and y` - if `x` is falsy the result is `x`, otherwise it is `y`

* `x or y` - if `x` is truthy the result is `x`, otherwise it is `y`

Either way the result is one of the original values, unchanged, whereas R's `&&` and <code>&#124;&#124;</code> always produce a single `TRUE`, `FALSE`, or `NA`.
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
1 and 2    # 1 is truthy -> 2
0 and 2    # 0 is falsy  -> 0
[] and [1] # [] is falsy -> []
```
:::

::: {.column width='50%'}
```{python}
0 or "default"     # 0 is falsy      -> "default"
"" or "default"    # "" is falsy     -> "default"
"abc" or "default" # "abc" is truthy -> "abc"
```
:::
::::


::: {.aside}
As mentioned last time, `0`, `""`, `None`, and empty containers are falsy, almost everything else is truthy.
:::

## Fallback values

::: {.small}
```{python}
#| eval: false
x or default
```
:::
is a common Python idiom for supplying a default when `x` is falsy. R 4.4 added the null coalescing operator `%||%` for the narrower case where `x` is `NULL`.

:::: {.columns .small}
::: {.column width='50%'}
```{python}
name = ""
name or "anonymous"
name = "Colin"
name or "anonymous"
```
:::

::: {.column width='50%'}
```{r}
name = NULL
name %||% "anonymous"
name = "Colin"
name %||% "anonymous"
```
:::
::::

::: {.aside}
`%||%` has long been available from `rlang` and the tidyverse.
:::


# Conditionals

## `if` and `else`

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = 3
if (x > 0) {
  print("x is positive")
}
```
```{r}
if (x > 0) {
  print("x is positive")
} else {
  print("x is not positive")
}
```
:::

::: {.column width='50%'}
```{python}
x = 3
if x > 0:
    print("x is positive")
```
```{python}
if x > 0:
    print("x is positive")
else:
    print("x is not positive")
```
:::
::::

. . .

R wraps the condition in `()` and the body in `{}`, while Python ends the condition with `:` and the body is the following *indented* block.

::: {.aside}
In R the `{}` are optional for a single-expression body, e.g. `if (x > 0) print("x is positive")`, but the [tidyverse style guide](https://style.tidyverse.org/syntax.html#if-statements) recommends braces for anything spanning multiple lines.
:::


## `else if` and `elif`

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = 0
if (x < 0) {
  print("x is negative")
} else if (x > 0) {
  print("x is positive")
} else {
  print("x is zero")
}
```
:::

::: {.column width='50%'}
```{python}
x = 0
if x < 0:
    print("x is negative")
elif x > 0:
    print("x is positive")
else:
    print("x is zero")
```
:::
::::

. . .

Conditions are checked in order and only the *first* true branch runs; the optional `else` branch runs if none are.


## Conditionals as expressions

::: {.medium}
R's `if` is an expression that returns the value of the evaluated branch, so it can be used directly in an assignment. Python's `if` is a *statement* that returns nothing; instead there is a separate *conditional expression*, `a if cond else b`.
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = 5
s = if (x %% 2 == 0) x / 2 else 3 * x + 1
s
```
:::

::: {.column width='50%'}
```{python}
x = 5
s = x / 2 if x % 2 == 0 else 3 * x + 1
s
```
:::
::::

. . .

Both are equivalent to assigning within each branch,

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
if (x %% 2 == 0) {
  s = x / 2
} else {
  s = 3 * x + 1
}
s
```
:::

::: {.column width='50%'}
```{python}
if x % 2 == 0:
    s = x / 2
else:
    s = 3 * x + 1

s
```
:::
::::


## Conditionals are *not* vectorized

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c(1, 3)
```
```{r}
#| error: true
if (x == 1) print("x is 1!")
```
```{r}
#| error: true
if (x == 3) print("x is 3!")
```
:::

::: {.column width='50%'}
```{python}
x = [1, 3]
```
```{python}
if x == 1:
    print("x is 1!")
else:
    print("x is not 1!")
```
```{python}
if [False, False]:
    print("non-empty lists are truthy")
```
:::
::::

. . .

Since R 4.2, `if` throws an error if the condition has length > 1 (older versions used the first value with a warning).

Python does not automatically apply the comparison element-wise - `x == 1` is simply `False`, and any non-empty list is truthy regardless of its contents.


## Collapsing logical vectors

Both languages provide `any()` and `all()` for reducing multiple logical values to a single one (base Python has no vectorized comparison, so the list of booleans is given directly).

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c(3, 4, 1)
x >= 2
any(x >= 2)
all(x >= 2)
```
:::

::: {.column width='50%'}
```{python}
x = [3, 4, 1]
any([True, True, False])
all([True, True, False])
```
:::
::::

. . .

:::: {.columns .small}
::: {.column width='50%'}
```{r}
if (any(x == 3)) print("x contains 3!")
```
:::

::: {.column width='50%'}
```{python}
if 3 in x: print("x contains 3!")
```
:::
::::


## Conditionals and truthiness

::: {.medium}
Python's conditionals accept any object and use its *truthiness*. R's `if` requires a single logical value, though it will coerce other values that `as.logical()` recognizes.
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
if (1) "yes" else "no"
if (0) "yes" else "no"
if ("TRUE") "yes" else "no"
```
```{r}
#| error: true
if ("abc") "yes" else "no"
```
```{r}
#| error: true
if (NULL) "yes" else "no"
```
:::

::: {.column width='50%'}
```{python}
"yes" if 1 else "no"
"yes" if 0 else "no"
"yes" if "abc" else "no"
"yes" if [] else "no"
"yes" if None else "no"
```
:::
::::


## Vectorized conditionals

R's `ifelse()` is the element-wise counterpart to `if` - given a logical vector it returns a vector of the same length built from the `yes` and `no` arguments.

::: {.small}
```{r}
x = c(-2, 0, 3)
ifelse(x > 0, "positive", "non-positive")
ifelse(x > 0, x, -x)
```
:::

. . .

Base Python has no equivalent; the idiomatic approach is a list comprehension with a conditional expression (next time).


## Conditionals and missing values

::: {.small}
* R - `NA` is sticky in comparisons and `if` errors on `NA` rather than guessing. `any()` and `all()` follow the <code>&#124;</code> and `&` rules, so `NA` is not always sticky.

* Python - `None` supports equality tests, but ordering it with a number raises `TypeError`. For `nan`, ordered comparisons and `==` are `False`; `!=` is `True`; and despite comparing equal to nothing, `nan` is truthy.

Check for missing values directly with `is.na()`, `x is None`, or `math.isnan()`.
:::

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
x = NA
x > 1
```
```{r}
#| error: true
if (x > 1) print("x is big")
```
```{r}
any(c(1, NA, 4) >= 3)
all(c(1, NA, 4) >= 1)
```
:::

::: {.column width='50%'}
```{python}
x = None
```
```{python}
#| error: true
if x > 1: print("x is big")
```
```{python}
if x is None: print("x is missing")
```
```{python}
(math.nan > 1, math.nan == math.nan, math.nan != math.nan)
bool(math.nan)
```
:::
::::


## Multi-way branching

R's `switch()` selects a branch based on a character (or integer) value, while Python 3.10+ has the `match` statement (which also supports much more general *structural pattern matching*).

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = "b"
switch(x,
  a = "apple",
  b = "banana",
  "unknown"
)
```
:::

::: {.column width='50%'}
```{python}
x = "b"
match x:
    case "a":
        print("apple")
    case "b":
        print("banana")
    case _:
        print("unknown")
```
:::
::::

::: {.aside}
`switch()` uses an unnamed final argument as the default (without one an unmatched value returns `NULL`, invisibly), `case _` is the wildcard for `match`.
:::


# Errors

## Signaling conditions

::: {.medium}
R has several ways of communicating with the user beyond `print()` / `cat()`, each of which is a *condition* that can be handled programmatically:

* `message()` - diagnostic messages (sent to stderr)

* `warning()` - something unexpected but not fatal, execution continues

* `stop()` - an error, execution halts

Python has rough equivalents in `print()` (or the `logging` module), `warnings.warn()`, and `raise` with an *exception* object.
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
message("Starting")
warning("Something unexpected")
```
:::

::: {.column width='50%'}
```{python}
import warnings
print("Starting")
warnings.warn("Something unexpected")
```
:::
::::

::: {.aside}
Errors stop a Quarto document from rendering unless the chunk sets `#| error: true` (which is how these slides show error output).
:::


## Raising errors

:::: {.columns .small}
::: {.column width='50%'}
```{r}
#| error: true
x = -1
if (x < 0) {
  stop("x must be non-negative")
}
```
:::

::: {.column width='50%'}
```{python}
#| error: true
x = -1
if x < 0:
    raise ValueError("x must be non-negative")
```
:::
::::

. . .

Both languages also have a shorthand for checking assumptions - R's `stopifnot()` and Python's `assert` statement:

:::: {.columns .small}
::: {.column width='50%'}
```{r}
#| error: true
stopifnot(x >= 0)
```
```{r}
#| error: true
stopifnot("x must be non-negative" = x >= 0)
```
:::

::: {.column width='50%'}
```{python}
#| error: true
assert x >= 0
```
```{python}
#| error: true
assert x >= 0, "x must be non-negative"
```
:::
::::

::: {.aside}
Python's `assert` statements are stripped when running with `python -O`, so use them for development-time checks rather than validating user input.
:::


## Errors vs exceptions

::: {.medium}
Python exceptions are objects with a class hierarchy - the class describes what went wrong and allows handling to be selective. R errors are, by default, all of the same class (`simpleError`) and are distinguished only by their message.
:::

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
#| error: true
"abc" + 1
```
```{r}
#| error: true
log("abc")
```
```{r}
#| error: true
sum("abc")
```
```{r}
#| error: true
undefined_var
```
:::

::: {.column width='50%'}
```{python}
#| error: true
"abc" + 1
```
```{python}
#| error: true
int("abc")
```
```{python}
#| error: true
[1, 2, 3][5]
```
```{python}
#| error: true
undefined_var
```
:::
::::

::: {.aside}
See the Python docs for the full [exception hierarchy](https://docs.python.org/3/library/exceptions.html#exception-hierarchy). Classed conditions are possible in R (e.g. via `rlang::abort()` / `cli::cli_abort()`) and are used throughout the tidyverse.
:::


## Handling errors

::: {.medium}
R's `try()` evaluates an expression and, instead of halting, returns a `try-error` object if an error occurs. Python's `try` / `except` block runs the `except` code only if a matching exception was raised in the `try` body.
:::

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = try(log("a"), silent = TRUE)
class(x)
cat(x)
inherits(x, "try-error")
```
:::

::: {.column width='50%'}
```{python}
try:
    x = math.log("a")
except TypeError as e:
    print("Caught:", e)
    x = math.nan
x
```
:::
::::


## `tryCatch()` and `try` / `except` / `finally`

::: {.medium}
R's `tryCatch()` lets you provide handler *functions* for specific condition classes (`error`, `warning`, `message`), while Python's `except` clauses are matched against the exception's class in order. Both support a `finally` block that always runs.
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
tryCatch(
  {
    log("a")
  },
  error = function(e) {
    cat("Error:", conditionMessage(e), "\n")
  },
  warning = function(w) {
    cat("Warning:", conditionMessage(w), "\n")
  },
  finally = {
    cat("Done\n")
  }
)
```
:::

::: {.column width='50%'}
```{python}
try:
    math.log("a")
except TypeError as e:
    print("Error:", e)
except ValueError as e:
    print("Value error:", e)
else:
    print("No error")
finally:
    print("Done")
```
:::
::::

::: {.aside}
Python's `else` clause runs only when no exception occurred. `tryCatch()` returns the value of the expression, or of the handler that was run.
:::


## Exercise 1

Without running the code, what do you expect the output (or error) to be for each of the listed values of `x`?

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
#| eval: false
if (x > 10 || x < -10) {
  stop("Input too big")
} else if (x %in% c(2, 3, 5, 7)) {
  cat("Input is prime!\n")
} else if (x %% 2 == 0) {
  cat("Input is even!\n")
} else if (x %% 2 == 1) {
  cat("Input is odd!\n")
}
```
```{r}
#| eval: false
x = 1
x = 3
x = -1
x = -3
x = 1:2
x = "0"
x = "zero"
```
:::

::: {.column width='50%'}
```{python}
#| eval: false
if x > 10 or x < -10:
    raise ValueError("Input too big")
elif x in [2, 3, 5, 7]:
    print("Input is prime!")
elif x % 2 == 0:
    print("Input is even!")
elif x % 2 == 1:
    print("Input is odd!")
```
```{python}
#| eval: false
x = 1
x = 3
x = -1
x = -3
x = [1, 2]
x = "0"
x = 2.5
```
:::
::::

```{r}
#| echo: false
countdown::countdown(minutes = 5)
```


# Loops

## `for` loops

R's `for` iterates over the elements of a vector (or list), while Python's `for` iterates over the elements of any *iterable* object (lists, tuples, strings, ranges, dictionaries, files, ...).

:::: {.columns .small}
::: {.column width='50%'}
```{r}
for (w in c("Hello", "world!")) {
  cat(w, ":", nchar(w), "\n")
}
```
```{r}
total = 0
for (v in c(1, 2, 3, 4)) {
  total = total + v
}
total
```
:::

::: {.column width='50%'}
```{python}
for w in ["Hello", "world!"]:
    print(w, ":", len(w))
```
```{python}
total = 0
for v in (1, 2, 3, 4):
    total += v
total
```
:::
::::



## Integer sequences

Loops over indices need a sequence of integers - R has `:`, `seq()`, `seq_len()`, and `seq_along()`, while Python has `range()`.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
1:5
seq(1, 10, by = 3)
seq_len(3)
seq_along(c("a", "b", "c"))
```
:::

::: {.column width='50%'}
```{python}
range(5)
list(range(5))
list(range(1, 11, 3))
list(range(5, 0, -1))
```
:::
::::

. . .

`range(start, stop, step)` excludes `stop` and is *lazy* - its size does not grow with the length of the range. For integer endpoints, R's `:` includes both; `seq()` includes `to` only when the step lands on it. R sequences behave like vectors but may be stored compactly internally.


## Looping over indices

The R idiom is `seq_along(x)`. Python's equivalent is `range(len(x))`, but `enumerate()` is preferred as it yields the index *and* the value together (as a tuple, which is unpacked into `i` and `v` below).

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c("a", "b", "c")
for (i in seq_along(x)) {
  cat(i, x[i], "\n")
}
```
:::

::: {.column width='50%'}
```{python}
x = ["a", "b", "c"]
for i in range(len(x)):
    print(i, x[i])
```
```{python}
for i, v in enumerate(x):
    print(i, v)
```
:::
::::

::: {.aside}
`enumerate(x, start=1)` can be used if 1-based indices are needed.
:::


## Avoid `1:length(x)`

The common R idiom `1:length(x)` fails for empty vectors since `1:0` is `c(1, 0)` - use `seq_along()` or `seq_len()` instead. Python's `range(len(x))` is safe since `range(0)` is empty.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = integer()
1:length(x)
seq_along(x)
```
```{r}
for (i in 1:length(x)) print(i)
for (i in seq_along(x)) print(i)
```
:::

::: {.column width='50%'}
```{python}
x = []
list(range(len(x)))
```
```{python}
for i in range(len(x)): print(i)
```
:::
::::


## Multiple sequences

Python's `zip()` iterates over multiple sequences together, stopping at the shortest. R has no direct equivalent - index with `seq_along()` instead (though vectorization usually makes this unnecessary).

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c(1, 2, 3)
y = c("a", "b", "c")
for (i in seq_along(x)) {
  cat(x[i], y[i], "\n")
}
```
:::

::: {.column width='50%'}
```{python}
x = [1, 2, 3]
y = ["a", "b", "c"]
for a, b in zip(x, y):
    print(a, b)
```
```{python}
list(zip([1, 2, 3, 4], "ab"))
```
:::
::::

::: {.aside}
`zip()` returns a lazy iterator (hence the `list()` to see its contents). See `itertools.zip_longest()` to iterate until the longest sequence is exhausted.
:::


## `while` loops

Repeat the body as long as the condition is `TRUE` / truthy - the condition is checked before each iteration, so the body may never run.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
i = 1
while (i < 100) {
  i = i * 2
}
i
```
:::

::: {.column width='50%'}
```{python}
i = 1
while i < 100:
    i *= 2
i
```
:::
::::

. . .

R also has `repeat`, which loops forever until a `break` - the Python idiom for this is `while True:`.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
i = 1
repeat {
  i = i * 2
  if (i >= 100) break
}
i
```
:::

::: {.column width='50%'}
```{python}
i = 1
while True:
    i *= 2
    if i >= 100:
        break
i
```
:::
::::


## `break` and `next` / `continue`

`break` exits the (innermost) loop entirely, while `next` in R and `continue` in Python skip the rest of the current iteration.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
for (i in 1:10) {
  if (i %% 3 == 0) next
  cat(i, "")
}
```
```{r}
for (i in 1:10) {
  if (i %% 3 == 0) break
  cat(i, "")
}
```
:::

::: {.column width='50%'}
```{python}
for i in range(1, 11):
    if i % 3 == 0:
        continue
    print(i, end=" ")
```
```{python}
for i in range(1, 11):
    if i % 3 == 0:
        break
    print(i, end=" ")
```
:::
::::


## Loop `else` and `pass`

Two Python-only constructs - loops can have an `else` clause that runs when the loop finishes *without* a `break`, and `pass` is a no-op placeholder for where a statement is syntactically required.

:::: {.columns .small}
::: {.column width='50%'}
```{python}
for n in range(2, 10):
    for x in range(2, n):
        if n % x == 0:
            print(n, "=", x, "*", n // x)
            break
    else:
        print(n, "is prime")
```
:::

::: {.column width='50%'}
```{python}
x = -3
if x < 0:
    pass
elif x % 2 == 0:
    print("x is even")
else:
    print("x is odd")
```
:::
::::

::: {.aside}
Loop `else` example from the [Python Tutorial](https://docs.python.org/3/tutorial/controlflow.html#else-clauses-on-loops). R has no loop `else`, and an empty block `{}` (or `NULL`) serves as a placeholder.
:::


## Building up results

:::: {.columns .small}
::: {.column width='50%'}
```{r}
res = c()
for (i in 1:5) {
  res = c(res, i^2)
}
res
```
:::

::: {.column width='50%'}
```{python}
res = []
for i in range(1, 6):
    res.append(i**2)
res
```
:::
::::

. . .

::: {.medium}
Growing an R vector with `c()` copies it every iteration (quadratic overall), so for longer loops preallocate (`numeric(n)`, `character(n)`, `vector("list", n)`) and assign by index. Python's `list.append()` is amortized constant time, so appending is idiomatic there.
:::

::: {.small}
```{r}
res = numeric(5)
for (i in seq_along(res)) {
  res[i] = i^2
}
res
```
:::

::: {.aside}
In practice most loops like these are replaced by vectorized operations, functional tools (`lapply()`, `purrr::map()`), or comprehensions - all coming up in the next few weeks.
:::


## Exercise 2

:::: {.columns .medium}
::: {.column width='40%'}
To the right are vectors containing all prime numbers between 2 and 100 and some values `x` we would like to check for primality.

Using *nested* loops, write code in both R and Python that prints only the values of `x` that are *not* prime - without using subsetting, `%in%`, or `in`.

In Python, try using the loop `else` clause; in R you will need a flag variable.
:::

::: {.column .xsmall width='60%'}
```{r}
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 
           29, 31, 37, 41, 43, 47, 53, 59, 61, 
           67, 71, 73, 79, 83, 89, 97)
x = c(3, 4, 12, 19, 23, 51, 61, 63, 78)
```
```{python}
primes = [ 2,  3,  5,  7, 11, 13, 17, 19, 23, 
          29, 31, 37, 41, 43, 47, 53, 59, 61, 
          67, 71, 73, 79, 83, 89, 97]
x = [3, 4, 12, 19, 23, 51, 61, 63, 78]
```
:::
::::

```{r}
#| echo: false
countdown::countdown(minutes = 5)
```


# Comparing R & Python

## Control flow summary

::: {.small}
| Construct              | R                                | Python                          |
|:-----------------------|:---------------------------------|:--------------------------------|
| conditional            | `if` / `else if` / `else`        | `if` / `elif` / `else`          |
| conditional expression | `if (cond) a else b`             | `a if cond else b`              |
| vectorized conditional | `ifelse()`                       | comprehension, `np.where()`     |
| multi-way branch       | `switch()`                       | `match` / `case`                |
| for loop               | `for (x in vec) {}`              | `for x in iterable:`            |
| while loop             | `while (cond) {}`                | `while cond:`                   |
| infinite loop          | `repeat {}`                      | `while True:`                   |
| skip iteration         | `next`                           | `continue`                      |
| exit loop              | `break`                          | `break`                         |
| integer sequences      | `:`, `seq_len()`, `seq_along()`  | `range()`                       |
| index + value          | `seq_along()` + `x[i]`           | `enumerate()`                   |
| multiple sequences     | `seq_along()` + indexing         | `zip()`                         |
| reduce logicals        | `any()`, `all()`                 | `any()`, `all()`                |
| blocks                 | `{ }`                            | `:` + indentation               |
:::


## Error handling summary

::: {.medium}
| Concept        | R                                   | Python                              |
|:---------------|:------------------------------------|:------------------------------------|
| message        | `message()`                         | `print()`, `logging`                |
| warning        | `warning()`                         | `warnings.warn()`                   |
| error          | `stop()`                            | `raise SomeError()`                 |
| assertion      | `stopifnot()`                       | `assert`                            |
| catch          | `try()`, `tryCatch()`               | `try` / `except`                    |
| always run     | `tryCatch(finally = )`, `on.exit()` | `finally`                           |
| error object   | condition (`simpleError`)           | exception (subclass of `Exception`) |
| error message  | `conditionMessage(e)`               | `str(e)`                            |
:::


## Takeaways

* R's comparison and logical operators (`==`, `&`, <code>&#124;</code>) are vectorized; `&&` / <code>&#124;&#124;</code> in R and `and` / `or` in Python are scalar and short-circuit.

* `if` in R needs exactly one `TRUE` / `FALSE` - use `any()`, `all()`, or `ifelse()` for vectors. Python's `if` tests the *truthiness* of any object.

* Loop syntax is nearly identical - the differences are blocks (`{}` vs indentation), `seq_along()` vs `range()` / `enumerate()`, and `next` vs `continue`.

* Errors are *conditions* in R (`stop()`, `tryCatch()`) and typed *exception* objects in Python (`raise`, `try` / `except`).
