---
title: "Data Structures<br/>in R & Python"
subtitle: "Lecture 06"
author: "Dr. Colin Rundel"
footer: "Sta 523 - Fall 2026"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
  warning: true
engine: knitr
---


```{r setup}
#| message: false
#| warning: false
#| include: false
options(
  width = 80
)
```

```{python py_setup}
#| include: false
import numpy as np
```


# Tuples

## Tuples

Python tuples are *heterogeneous*, *ordered*, **immutable** containers - lists whose contents cannot be changed. R has no tuple type, a list (or an atomic vector) is used instead.

:::: {.columns .small}
::: {.column width='50%'}
```{python}
(1, 2, 3)
(1, True, "abc")
```
```{python}
type( (1) )
type( (1,) )
```
:::

::: {.column width='50%' .fragment}
```{python}
#| error: true
x = (1, 2, 3)
x[2] = 5
```
```{python}
x = (1, [2, 3])
x[1].append(4)
x
```
:::
::::

::: {.aside}
It is the *comma* that creates a tuple, not the parentheses - `(1)` is just the number 1 and a one element tuple must be written `(1,)`. Immutability is *shallow*, a mutable object stored in a tuple can still be modified in place.
:::


## Unpacking

Assigning a sequence to several names at once *unpacks* it - this is how Python functions return multiple values. A name prefixed with `*` collects any remaining values into a list.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
def minmax(x):
    return min(x), max(x)
minmax([3, 1, 2])
lo, hi = minmax([3, 1, 2])
print(lo, hi)
```
```{python}
x, y = 1, 2
x, y = y, x
x, y
```
:::

::: {.column width='50%' .fragment}
```{python}
first, *rest = range(6)
first
rest
```
```{python}
pairs = [("a", 1), ("b", 2)]
for i, (k, v) in enumerate(pairs):
    print(i, k, v)
```
:::
::::

::: {.aside}
The number of names must match the length of the sequence (`x, y = [1, 2, 3]` is a `ValueError`) unless one of them is starred. The same unpacking happens in `for` loops with `enumerate()`, `zip()`, and `dict.items()`.
:::


# Dictionaries

## Dictionaries

Python `dict`s are *mutable* containers of **key: value** pairs, designed for the efficient lookup of a value by its key. The closest R analog is a named list (or a named atomic vector when the values are homogeneous).

:::: {.columns .small}
::: {.column width='50%'}
```{python}
{"abc": 123, "def": 456}
dict([("abc", 123), ("def", 456)])
dict(abc=123, hello=456)
dict(zip(["abc", "def"], [1, 2]))
```
:::

::: {.column width='50%'}
```{r}
list(abc = 123, def = 456)
```
```{r}
c(abc = 123, def = 456)
```
:::
::::

::: {.aside}
Since Python 3.7 dictionaries preserve insertion order. The keyword form of `dict()` only works for keys that are valid Python names (`def` is a reserved word).
:::


## Keys must be hashable

Dictionary keys and set elements must be **hashable**: their hash must remain stable during their lifetime, and objects that compare equal must have equal hashes. Dictionary values can be anything.

::: {.small}
| Hashable                                      | Unhashable                                  |
|:----------------------------------------------|:--------------------------------------------|
| `int`, `float`, `bool`, `complex`               | `list`                                      |
| `str`, `bytes`, `range`                        | `dict`                                      |
| `None`, `frozenset`                            | `set`                                       |
| `tuple` if all its elements are hashable       | `tuple` containing an unhashable element     |
:::

. . .

Lists are unhashable, so neither a list nor a tuple containing a list can be a key:

:::: {.columns .small}
::: {.column width='50%'}
```{python}
#| error: true
{[1]: "bad"}
```
:::

::: {.column width='50%'}
```{python}
#| error: true
{(1, [2]): "bad"}
```
:::
::::

::: {.aside}
`hash()` returns an integer used to locate an entry in a hash table. Hashability is distinct from immutability: instances of user-defined classes are hashable by identity by default, even when their attributes can change. Defining `__eq__` in a class without defining `__hash__` makes its instances unhashable.
:::


## Lookup

`[]` raises a `KeyError` for a missing key, `in` tests for a key and `.get()` supplies a default.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
#| error: true
x = {1: "abc", "y": "hello", (1, 1): 3.14}
x[1]
x[(1, 1)]
x["def"]
```
:::

::: {.column width='50%' .fragment}
```{python}
"y" in x
"hello" in x
print( x.get("def") )
x.get("def", 0)
```
:::
::::


## Insert, replace, remove

Assigning to a key inserts or replaces, `del` and `.pop()` remove. Note that assigning `None` does *not* remove a key, whereas in R assigning `NULL` does.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
x = {1: "abc", "y": "hello"}
x["def"] = -1
x["y"] = "goodbye"
x
```
```{python}
del x[1]
x.pop("def")
x["y"] = None
x
```
:::

::: {.column width='50%'}
```{r}
y = list(abc = 123, def = 456)
y$ghi = -1
y$def = "goodbye"
str(y)
```
```{r}
y$abc = NULL
str(y)
```
:::
::::


## Methods & iteration

Iterating over a `dict` yields its *keys* - use `.items()` to get keys and values together.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
x = {1: "abc", "y": "hello"}
x.keys()
x.values()
x.items()
```
```{python}
for k, v in x.items():
    print(k, "->", v)
```
:::

::: {.column width='50%' .fragment}
```{python}
x | {"y": "goodbye", "w": 0}
```
```{python}
x.update({"y": "goodbye", "w": 0})
x
```
:::
::::

::: {.aside}
`.keys()`, `.values()`, and `.items()` return dynamic *views* of the dictionary rather than copies - see the [docs](https://docs.python.org/3/library/stdtypes.html#dictionary-view-objects). `|` returns a new merged dictionary (right side wins), `.update()` merges in place.
:::


# Environments

## Environments

An environment is a set of name-object *bindings* - the workspace is one, each package is one, and every function call creates a fresh one for its local variables. A lookup that fails in an environment continues in its *parent*.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = 1
f = function() {
  y = 2
  environment()
}
e = f()
ls(e)
```
```{r}
parent.env(e)
get("x", envir = e)
```
:::

::: {.column width='50%' .fragment}
```{r}
g = function() {
  z = 3
  z
}
g()
exists("z")
```
:::
::::

::: {.aside}
`globalenv()` is the workspace and the chain of parents ends at `emptyenv()`. Looking up through the parents is *lexical scoping* - `f()` finds `x` in the global environment, and `z` is gone once `g()` returns. More on this with functional programming.
:::


## Environments as hash maps

Bindings are stored in a hash table, so an environment makes a good `dict` - use `parent = emptyenv()` so that lookups do not fall through to the workspace.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
h = new.env(parent = emptyenv())
h$a = 1
h[["b"]] = 2
```
```{r}
ls(h)
h$a
exists("x", envir = h)
rm("a", envir = h)
ls(h)
```
:::

::: {.column width='50%' .fragment}
```{r}
h2 = h
h2$d = 4
ls(h)
```
```{r}
#| error: true
h[1]
```
:::
::::

::: {.aside}
Modifying `h2` changed `h` - environments have *reference* semantics, both names refer to the same object exactly like two Python names bound to one `dict`. `utils::hashtab()` (R 4.2+) provides a hash table with arbitrary keys.
:::


## Exercise 1

Using the record from last lecture, complete these tasks in Python:

:::: {.columns}
::: {.column width='55%'}
::: {.xsmall}
```python
person = {
  "firstName": "John",
  "lastName": "Smith",
  "age": 25,
  "address": {
    "streetAddress": "21 2nd Street",
    "city": "New York",
    "state": "NY",
    "postalCode": 10021
  },
  "phoneNumber": [
    { "type": "home",
      "number": "212 555-1239" },
    { "type": "fax",
      "number": "646 555-4567" }
  ]
}
```
:::
:::

::: {.column width='45%'}
::: {.small}
* Extract the fax number.

* Add a `"mobile"` phone number and remove `age`.

* Write `phone(person, phone_type)` to return the requested number, or `None` if absent.

* **Extension:** Write the equivalent function in R.
:::
:::
::::

```{r}
#| echo: false
countdown::countdown(minutes = 5)
```

::: {.aside}
JSON uses `true`, `false`, and `null` where Python uses `True`, `False`, and `None`. Use `json.loads()` to parse JSON text.
:::


# Sets

## Sets

A `set` is a *mutable*, *unordered* collection of **unique** hashable elements - there are no positions, so the primary operation is membership testing with `in`.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
x = {1, 2, 3, 4, 1, 2}
x
set("mississippi")
```
```{python}
#| error: true
3 in x
x[0]
```
:::

::: {.column width='50%' .fragment}
```{python}
x.add(9)
x.discard(8)
x.update([7, 8])
x
```
```{python}
#| error: true
x.remove(6)
```
```{python}
#| error: true
{1, 2, [1, 2]}
```
:::
::::

::: {.aside}
`{}` creates an empty *dictionary*, an empty set must be written `set()`. `remove()` raises a `KeyError` if the element is absent while `discard()` does not, and iteration order is not guaranteed (use `sorted(x)` for a stable order).
:::


## Set operations

Sets support the usual mathematical operations. R has no set type - any vector can be treated as one via `unique()` and the set functions.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
x = {1, 2, 3}
y = {2, 3, 4}
x | y
x & y
x - y
1 in x
```
:::

::: {.column width='50%'}
```{r}
x = c(1, 2, 3, 1)
y = c(2, 3, 4)
union(x, y)
intersect(x, y)
setdiff(x, y)
1 %in% x
```
:::
::::

::: {.aside}
Python's set methods (`x.union(y)`, `x.intersection(y)`, ...) accept any iterable, the operators require sets on both sides. R's set functions return unique values in order of first appearance.
:::


# Mutability & copies

## Nested containers

In Lecture 2 we saw that `.copy()` is *shallow*, and `copy.deepcopy()` copies recursively. The same sharing problem occurs when building nested structures by repetition,

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
grid = [[0] * 3] * 3
grid[0][0] = 1
grid
```
```{python}
grid = [[0] * 3 for _ in range(3)]
grid[0][0] = 1
grid
```
:::

::: {.column width='50%' .fragment}
```{python}
import copy

x = {"a": [1, 2], "b": [3, 4]}
y = x.copy()
z = copy.deepcopy(x)

x["a"].append(99)
y
z
```
:::
::::

::: {.aside}
`[[0] * 3] * 3` creates *one* inner list and references it three times. The comprehension evaluates `[0] * 3` three separate times.
:::


## Mutable default arguments

Default values are evaluated *once*, when the function is defined, so a mutable default is shared by every call that does not supply the argument. Use `None` as the default and create the object inside the function instead.

:::: {.columns .small}
::: {.column width='50%'}
```{python}
def f(x=[]):
    x.append(1)
    return x

f()
f()
```
:::

::: {.column width='50%' .fragment}
```{python}
def f(x=None):
    if x is None:
        x = []
    x.append(1)
    return x

f()
f()
```
:::
::::


# Algorithms & data structures


## Big-O notation

::: {.medium}
Big-O notation describes the *complexity* of an algorithm - how the time (or memory) required grows with the size of the input $n$. Only the fastest growing term matters (constant factors and lower order terms are dropped), so two algorithms with the same Big-O can still differ greatly in practice.

Since performance depends on the data we usually quote *average* or *worst case* complexity, and *amortized* complexity when an occasional expensive step is paid for by many cheap ones (e.g. growing a list).
:::

::: {.small .center}
| Complexity  | Big-O           |
|-------------|-----------------|
| Constant    | O($1$)          |
| Logarithmic | O($\log n$)     |
| Linear      | O($n$)          |
| Quasilinear | O($n \log n$)   |
| Quadratic   | O($n^2$)        |
| Exponential | O($C^n$)        |
:::


## Under the hood

The containers in both languages are built from a handful of basic structures - knowing which is which explains what each container is good, and bad, at.

::: {.small}
| Structure              | Layout                                              | R                                  | Python              |
|:-----------------------|:----------------------------------------------------|:-----------------------------------|:--------------------|
| Array                  | contiguous block of same-sized values               | materialized numeric vector        | contiguous numeric `ndarray` |
| Array of pointers      | contiguous block of *references* to objects         | list                               | `list`, `tuple`     |
| Linked list            | nodes that each point to their neighbors            | pairlist                           | `deque` (linked blocks) |
| Hash table             | array indexed by `hash(key)`                        | environment                        | `dict`, `set`       |
:::

::: {.aside}
See [R Internals](https://cran.r-project.org/doc/manuals/r-release/R-ints.html) and [CPython Internals](https://github.com/zpoint/CPython-Internals) for the gory details.
:::



## Vectors

Materialized numeric vectors in R and contiguous numeric NumPy arrays store fixed-size values together, giving O(1) element access. NumPy views can have gaps or reversed strides.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
object.size(numeric(1e6))
object.size(integer(1e6))
```
:::

::: {.column width='50%'}
```{python}
np.zeros(1_000_000).nbytes
np.zeros(1_000_000, dtype="int64").nbytes
np.zeros(1_000_000, dtype="int32").nbytes
```
:::
::::

::: {.aside}
R doubles use 8 bytes per element; integers use 4. `object.size()` includes object overhead; NumPy's `.nbytes` counts only element storage.
:::


## Growing a vector

Repeatedly appending with `c(x, i)` copies the existing elements each time, giving O($n^2$) work overall.


::: {.xsmall}
```{r}
n = 5e4
x = c()
system.time(
  for (i in seq_len(n)) x = c(x, i)
)
```
:::

Indexed growth can reuse spare capacity, but still needs occasional copying. Preallocation avoids repeated growth.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = numeric()
system.time(
  for (i in seq_len(n)) x[i] = i
)
```
:::

::: {.column width='50%' .fragment}
```{r}
x = numeric(n)
system.time(
  for (i in seq_len(n)) x[i] = i
)
```
:::
::::


## Generic vectors

R and Python lists store references to objects. Elements can have different types and sizes, while indexing remains O(1). References and object headers add memory overhead.

::: {.small}
```text
Numeric vector: values stored together

[ 1.0 | 2.0 | 3.0 ]


List: references stored together

[  •  |  •  |  •  ]
   ↓     ↓     ↓
   1   "abc"  TRUE
```
:::

::: {.notes}
Every R object is a `SEXP`, a pointer to a struct holding

* a header with the type (`SEXPTYPE`), flags, and garbage collector bookkeeping

* a pointer to the object's attributes (a pairlist)

* the data
  - for an atomic vector (`REALSXP`, `INTSXP`, ...) the length followed by a block of values
  - for a list (`VECSXP`) the length followed by an array of `SEXP`s

Every Python value is a `PyObject`, a struct holding

* a reference count (`ob_refcnt`) and a pointer to its type (`ob_type`)

* type specific data 
  - a `float` stores its value

  - a `list` stores its length, its allocated capacity, and a pointer to an array of `PyObject *`
:::

::: {.aside}
CPython lists reserve extra slots, making appends amortized O(1). Resizing may copy the pointer array. See [R Internals](https://cran.r-project.org/doc/manuals/r-release/R-ints.html#SEXPs) and the [CPython docs](https://docs.python.org/3/c-api/structures.html) for the full structs.
:::


## Array

<iframe data-src="https://visualgo.net/en/array" width="100%" height="450px" style="border:1px solid;border-radius: 5px;" data-external="1">
</iframe>

::: {.aside}
<https://visualgo.net/en/array> - Materialized numeric R vectors and contiguous numeric NumPy arrays store same-typed values together. A Python `list` is a contiguous array of *pointers* to objects, which is why it can be heterogeneous and why it uses far more memory.
:::


## deques

A `deque` (double-ended queue) supports O(1) additions and removals at either end. CPython stores it as a doubly linked chain of fixed-size blocks. Indexing is O(1) near either end and O(n) in the middle.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
from collections import deque
x = deque(range(3))
x.appendleft(-1)
x
```
```{python}
x.popleft()
x
```
:::

::: {.column width='50%' .fragment}
```{python}
x = deque(range(3), maxlen=4)
x.append(10)
x.append(11)
x
```
```{python}
x.appendleft(-1)
x
```
:::
::::

::: {.aside}
With `maxlen` set, adding to a full deque drops an element from the opposite end - a rolling window of the last $n$ items. `list.pop(0)` and `list.insert(0, v)` have to shift every remaining element.
:::


## Linked list

<iframe data-src="https://visualgo.net/en/list" width="100%" height="450px" style="border:1px solid;border-radius: 5px;" data-external="1">
</iframe>

::: {.aside}
<https://visualgo.net/en/list> - CPython's `deque` links blocks of elements. R uses *pairlists* (linked lists) internally for function arguments and attributes.
:::


## Hash tables in practice

Hash-table membership is O(1) on average; searching a list is O(n). These timings illustrate the practical difference:

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
import time
big_list = list(range(100_000))
big_set = set(big_list)
queries = range(0, 100_000, 100)

start = time.perf_counter()
res = [q in big_list for q in queries]
print(time.perf_counter() - start)

start = time.perf_counter()
res = [q in big_set for q in queries]
print(time.perf_counter() - start)
```
:::

::: {.column width='50%'}
```{r}
big = sample(1e5)
queries = sample(1e5, 1e3)

system.time({
  for (q in queries) any(big == q)
})
system.time({
  queries %in% big
})
```
:::
::::

::: {.aside}
Python's set is built before timing; R's `%in%` includes lookup preparation. To study scaling, vary the container size while keeping the number of queries fixed. Use `timeit` (Python) or `bench::mark()` (R) for repeated timings.
:::


## Hash table

<iframe data-src="https://visualgo.net/en/hashtable" width="100%" height="450px" style="border:1px solid;border-radius: 5px;" data-external="1">
</iframe>

::: {.aside}
<https://visualgo.net/en/hashtable> - Python's `dict` and `set` are hash tables, as are R environments. R's `match()`, `%in%`, `unique()`, and `duplicated()` build a hash table of one argument, which is why they are fast. Equal keys must hash equally, so `1`, `1.0`, and `True` are the same `dict` key.
:::


## Sequence operation costs

::: {.small}
| Operation                 | `list`                         | `deque`                        |
|:--------------------------|:-------------------------------|:-------------------------------|
| Copy                      | `.copy()`: O(n)                | `.copy()`: O(n)                |
| Get / replace by index    | `s[i]` / `s[i] = v`: O(1)      | `s[i]` / `s[i] = v`: O(n)      |
| Append at end             | `.append(v)`: O(1)\*           | `.append(v)`: O(1)             |
| Remove from end           | `.pop()`: O(1)\*               | `.pop()`: O(1)                 |
| Insert at front           | `.insert(0, v)`: O(n)          | `.appendleft(v)`: O(1)         |
| Remove from front         | `.pop(0)`: O(n)                | `.popleft()`: O(1)             |
| Insert / delete in middle | O(n)                           | O(n)                           |
| Membership                | `v in s`: O(n)                 | `v in s`: O(n)                 |
:::

::: {.aside}
CPython costs for n elements. \*List append and end-pop costs are amortized. Deque indexing is O(1) near either end and O(n) in the middle. See <https://wiki.python.org/moin/TimeComplexity>.
:::


## Dictionary and set operation costs

::: {.small}
| Operation       | `dict`                | `set`                 |
|:----------------|:----------------------|:----------------------|
| Copy            | `.copy()`: O(n)       | `.copy()`: O(n)       |
| Lookup value    | `d[k]`: O(1)          | —                     |
| Insert / update | `d[k] = v`: O(1)      | `.add(v)`: O(1)       |
| Remove          | `.pop(k)`: O(1)       | `.remove(v)`: O(1)    |
| Membership      | `k in d`: O(1)        | `v in s`: O(1)        |
:::

::: {.aside}
Hash-table operations are O(1) on average, assuming constant-cost hashing and equality; individual operations can take O(n). Insertions are amortized over occasional table resizing. Dictionary membership tests keys.
:::


## Exercise 2

Choose two scenarios. For each, suggest a data structure in Python and R. Explain which operations your choice makes efficient, and state any assumptions.

::: {.small}
* A collection of 100 integers used repeatedly in elementwise arithmetic.

* A queue (first in, first out) of customer records.

* A stack (first in, last out) of customer records.

* A count of word occurrences within a document.

* The heights of the bars in a histogram with even bin widths.

* Checking whether each of a million words is in a list of 50 stop words.

* The last 10 lines read from a very large log file.
:::

```{r}
#| echo: false
countdown::countdown(minutes = 3)
```


# Comparing R & Python

## Container summary

::: {.small}
| Python                     | Properties                              | R                                           |
|:---------------------------|:----------------------------------------|:--------------------------------------------|
| `list`                     | ordered, mutable, heterogeneous         | `list()`                                    |
| `tuple`                    | ordered, immutable, heterogeneous       | `list()` or atomic vector (copy on modify)  |
| `dict`                     | key lookup, mutable, insertion ordered  | named `list()`, `new.env()` for hashing     |
| `set`                      | unique, unordered, mutable              | vector + `unique()`, `union()`, `%in%`, ... |
| `deque`                    | fast at both ends                       | none (use a list or vector)                 |
| `range`                    | lazy integer sequence, immutable        | `seq_len()`, `:` (often stored compactly via ALTREP) |
| NumPy `ndarray`            | homogeneous, vectorized                 | atomic vector, matrix, array                |
:::


## Takeaways

* **Tuples** are immutable sequences—use them for fixed groups of values and unpacking.

* **Dictionaries** map hashable keys to values, with O(1) lookup on average. R offers named lists and hash-based environments.

* **Sets** provide O(1) membership on average and set algebra. In R, use `%in%`, `union()`, `intersect()`, and `setdiff()`.

* **Shallow copies** share nested objects. Use `deepcopy()` when nested objects need independent mutation.

* **Mutable defaults** are shared between calls. Use `None` and create the object inside the function.

* Choose containers for their operations: deques for queues, sets for membership, and preallocated R vectors when the size is known. Measure when performance matters.
