---
title: "Subsetting &<br/>Vectorization"
subtitle: "Lecture 04"
author: "Dr. Colin Rundel"
footer: "Sta 523 - Fall 2026"
format:
  revealjs:
    theme: slides.scss
    transition: fade
    slide-number: true
    self-contained: true
execute:
  echo: true
  warning: true
engine: knitr
---


```{r setup}
#| message: false
#| warning: false
#| include: false
options(
  width = 80
)
```

```{python py_setup}
#| echo: false
import numpy as np
```


# Vectorization in R

## Vectorized operations in R

*Vectorized* code applies one operation to many values without writing an explicit loop. R's arithmetic, comparison, and element-wise logical operators are vectorized.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = 1:5
x^2
sqrt(x)
x + 10
```
:::

::: {.column width='50%'}
```{r}
x > 3
x %% 2 == 0
(x > 1) & (x < 5)
```
:::
::::

::: {.aside}
The reason we care about vectorization is that it can be orders of magnitude faster than an explicit loop.
:::

## R recycles by length

When vector lengths differ, R repeats the shorter vector to match the longer one.

::: {.xsmall}
```{r}
x = 1:6
```
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x + 10
x + c(10, 100)
```
:::

::: {.column width='50%'}
```{r}
x + c(10, 100, 1000)
x + c(10, 100, 1000, 10000)
```
:::
::::

. . .

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x > 3
x > c(2, 4)
```
:::

::: {.column width='50%'}
```{r}
x == c(1, 2, 3)
c(TRUE, FALSE) & c(TRUE, TRUE, TRUE)
```
:::
::::

::: {.aside}
R warns only when the longer length is not a multiple of the shorter length.
:::

## Zero-length recycling

Zero-length vectors contain no values; `NULL` also has length zero. For vectorized operations the presence of a length 0 vector means the result will also have length 0.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
numeric()
integer()
logical()
character()
NULL
length(NULL)
```
:::

::: {.column width='50%'}
```{r}
x = 1:4
```
```{r}
x + numeric()
x > numeric()
c(TRUE, FALSE) & logical()
x + NULL
x == NULL
```
:::
::::



# Vectorization in Python

## Python lists are not vectorized

Python's list operators manipulate the container, not the values. Element-wise arithmetic and comparisons instead require explicit iteration: a `for` loop or, more idiomatically, a comprehension.

::: {.small}
```{python}
x = [1, 2, 3]
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
x * 2
x + x
```
:::

::: {.column width='50%'}
```{python}
#| error: true
x + 2
```
```{python}
#| error: true
x > 2
```
```{python}
x + [2]
x > [2]
```
:::
::::

::: {.aside}
`*` repeats a list and `+` concatenates two lists
:::

## List comprehensions

A comprehension combines an expression with a `for` clause to construct a list. A trailing `if` filters values; a conditional expression transforms every value.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
result = []
for x in range(6):
    result.append(x**2)
result
[x**2 for x in range(6)]
```
:::

::: {.column .fragment width='50%'}
```{python}
[x**2 for x in range(10)
 if x % 2 == 0]
[x if x % 2 == 0 else -1
 for x in range(10)]
```
:::
::::

. . .

Comprehensions can be nested. Additional `for` clauses act like nested loops and produce one flat list, while nested comprehensions build a list of lists.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
[(i, j)
 for i in range(3)
 for j in range(2)]
```
:::

::: {.column width='50%'}
```{python}
[[i * j for j in range(4)]
 for i in range(3)]
```
:::
::::


## NumPy

[NumPy](https://numpy.org/) is a third-party Python package for numerical computing. Its main data structure is the homogeneous, multidimensional `ndarray`, with operations designed to work efficiently across entire arrays (i.e. vectorization).

. . .

Installation happens once per project (e.g. `uv add numpy`). Before using NumPy in a Python session or script, import it; `np` is the conventional short alias, so functions are called as `np.array()`, `np.arange()`, `np.sqrt()`, etc.

::: {.small}
```{python}
import numpy as np
np.__version__
```
:::

## Vectorized operations with NumPy

NumPy arrays support the same style of vectorized arithmetic, comparison, and logical operations as R's atomic vectors, with similar performance compared to explicit for loops.

:::: {.columns .small}
::: {.column width='50%'}
```{python}
x = np.arange(1, 6)
x**2
np.sqrt(x)
x + 10
```
:::

::: {.column width='50%'}
```{python}
x > 3
x % 2 == 0
(x > 1) & (x < 5)
```
:::
::::

::: {.aside}
Careful use of parentheses can be needed to make sure the correct operator(s) are used.
:::


## Vectorized conditional choice

`np.where()` is NumPy's equivalent to R's `ifelse()`, both construct a new vector by choosing between two values element-wise based on a condition.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = 0:7
ifelse(x %% 2 == 0, x, -1)
ifelse(x > 5, "big", "small")
```
:::

::: {.column width='50%'}
```{python}
x = np.arange(8)
np.where(x % 2 == 0, x, -1)
np.where(x > 5, "big", "small")
```
:::
::::

. . .

With a single argument, `np.where()` instead returns the positions where the condition is true; the R equivalent is `which()`.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
which(x > 5)
```
:::

::: {.column width='50%'}
```{python}
np.where(x > 5)
```
:::
::::


# Subsetting

## Indexing conventions

R uses 1-based indexing; Python uses 0-based indexing when subsetting by positive integers.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c("a", "b", "c", "d", "e")
```
```{r}
x[1]
x[5]
```
:::

::: {.column width='50%'}
```{python}
x = ["a", "b", "c", "d", "e"]
```

```{python}
x[0]
x[4]
```
:::
::::

. . .

In R, a negative index *excludes* a position. In Python, it counts backward from the end (`-1` is the last element).

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x[-1]
```
:::

::: {.column width='50%'}
```{python}
x[-1]
```
:::
::::


## Python slices

A slice has the form `start:stop:step`; `start` is included and `stop` is excluded. Any component may be omitted.

::: {.xsmall}
```{python}
x = list("abcdefgh"); x
```
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
x[1:5]
x[:4]
x[4:]
x[:]
```
:::

::: {.column width='50%'}
```{python}
x[1:7:2]
x[::2]
x[::-1]
x[-3:]
```
:::
::::


## Selecting multiple positions

R accepts a vector of positions inside `[ ]`. A base Python list does not: a slice can select a regular range, but arbitrary positions require a comprehension.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = letters[1:5]
```
```{r}
x[c(1, 3, 5)]
x[c(4, 2, 2)]
```
```{r}
#| error: true
x[c(1, -1)]
```
:::

::: {.column width='50%'}
```{python}
x = list("abcde")
```
```{python}
#| error: true
x[[0, 2, 4]]
```
```{python}
[x[i] for i in [0, 2, 4]]
[x[i] for i in [3, 1, 1]]
[x[i] for i in [0, -1]]
```
:::
::::

::: {.aside}
Both languages preserve the requested order and repeated positions. R cannot mix positive and negative indexes in one subset; Python can because each integer independently identifies a position.
:::


## Integer array indexing

Unlike base Python lists, NumPy arrays accept a list or integer array of positions. As with list indexing, negative positions count backward from the end and can be mixed with positive positions.

::: {.small}
```{python}
x = np.arange(10, 60, 10); x
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
x[[0, 2, 4]]
x[[3, 1, 1]]
```
:::

::: {.column width='50%'}
```{python}
x[np.array([0, 2, 4])]
x[[0, -1]]
```
:::
::::

::: {.aside}
Selecting with an integer array or list is called *advanced indexing*; selecting with a scalar or slice is *basic indexing*. The distinction will matter shortly.
:::


## Out-of-bounds positions

Out-of-bound indexing fails in Python, while R returns a typed missing value for a position that does not exist. Python slices clip each endpoint to the nearest valid boundary.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = c(10, 20, 30)
```
```{r}
x[4]
x[c(1, 4)]
x[-4]
```
:::

::: {.column width='50%'}
```{python}
x = [10, 20, 30]
```
```{python}
#| error: true
x[4]
```
```{python}
x[:4]
x[-4:]
x[4:]
x[:-4]
```
:::
::::


## R - Four additional index types

We have already used positive integers to select positions and negative integers to exclude positions. R's `[` operator supports four additional index forms:

<br/>


* logicals - select by`TRUE` positions

* characters - select by names

* zero - select nothing

* empty index - select everything


## Logical indexes

Logical subsetting keeps values corresponding to `TRUE`. 

Most logical indexes are created by vectorized logical expressions, which return one logical value for each element, often called a *mask*.


::: {.xsmall}
```{r}
x = c(10, 15, 20, 25, 30)
```
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x %% 20 == 0
x[x %% 20 == 0]
```
:::

::: {.column width='50%'}
```{r}
keep = x > 18
keep
x[keep]
```
:::
::::

. . .

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
(x > 10) & (x < 30)
x[(x > 10) & (x < 30)]
```
:::

::: {.column width='50%'}
```{r}
x[c(TRUE, NA, FALSE, TRUE, FALSE)]
```
:::
::::


::: {.aside}
Logical indexes are recycled to the length of `x`; an `NA` in the index produces an `NA` in the result.
:::


## NumPy & Boolean index

::: {.medium}
NumPy also supports using Boolean arrays for subsetting - positions where the index is `True` are kept and those with `False` are discarded. Boolean ndarrays, lists of Booleans, and the masks produced by vectorized comparisons all work.
:::

::: {.xsmall}
```{python}
x = np.arange(6)
```
:::


::: {.columns .xsmall}
::: {.column}
```{python}
x[np.array([True, False, True, False, True, False])]
x[[True, False, True, False, True, False]]
```
:::

::: {.column}
```{python}
x[x % 2 == 0]
x[x > 2]
```
:::
:::



. . .

::: {.medium}
However, while R recycles logical vectors, NumPy requires a Boolean index to match the dimension it selects.
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = 0:5
```
```{r}
x[c(TRUE, FALSE)]
```
:::

::: {.column width='50%'}
```{python}
#| error: true
x[np.array([True, False])]
```
```{python}
#| error: true
x[[True]]
```
:::
::::


## Boolean expressions with NumPy

NumPy overloads `&`, `|`, and `~` as element-wise logical operators. 

Parenthesize every comparison because these operators have unexpected precedence.

::: {.small}
```{python}
x = np.arange(10)
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
(x > 2) & (x < 7)
x[(x > 2) & (x < 7)]
```
:::

::: {.column width='50%'}
```{python}
(x < 2) | (x > 7)
x[~(x % 2 == 0)]
```
:::
::::

::: {.aside}
Do not use `and`, `or`, or `not` with whole NumPy arrays.
:::


## Character indexes

Names allow values to be selected independently of their positions.

::: {.xsmall}
```{r}
x = c(a = 10, b = 20, c = 30)
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x["a"]
x[c("c", "a", "a")]
```
:::

::: {.column width='50%'}
```{r}
x["d"]
x[c("a", "d")]
unname(x["a"])
```
:::
::::

::: {.aside}
Character subsetting returns the first value with each requested name. Names do not need to be unique.
:::


## Empty and zero indexes

In R an empty index selects everything, a zero index selects nothing. This differs from Python, where `0` is the first position and `:` is used to select everything.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = c(10, 20, 30)
```
```{r}
x[]
x[0]
x[NULL]
x[c(0, 2, 0)]
```
:::

::: {.column width='50%'}
```{python}
x = [10, 20, 30]
```
```{python}
x[:]
x[0]
x[0:0]
```
:::
::::

::: {.aside}
In R, zeros may appear alongside positive or negative indexes and are ignored. `NULL` and an empty integer vector also select no values.
:::


## Subsetting and assignment

Subset syntax can appear on the left side of an assignment in all three cases. A base Python list supports assignment to a single position or a slice, while R and NumPy also allow assignment via logical / Boolean and integer indexes.

:::: {.columns .xsmall}
::: {.column width='33%'}
```{r}
x = c(1, 4, 7, 9)
x[2] = 2
x[x %% 2 != 0] = -1
x
```
:::

::: {.column width='33%'}
```{python}
x = [1, 4, 7, 9]
x[1] = 2
x[2:4] = [0, 0, 0]
x
```
:::

::: {.column width='33%'}
```{python}
x = np.array([1, 4, 7, 9])
x[1] = 2
x[x % 2 != 0] = -1
x
```
:::
::::

::: {.aside}
A slice assignment can change a list's length. NumPy arrays have fixed size and type - assignment can replace values, but cannot insert elements or change the array's `dtype`.
:::


## NumPy views and copies

Basic slicing usually returns a *view* that shares data with the original array. Advanced indexing returns a *copy*.

:::: {.columns .small}
::: {.column width='50%'}
```{python}
x = np.arange(6)
y = x[1:4]
y[0] = 99
x
np.shares_memory(x, y)
```
:::

::: {.column width='50%'}
```{python}
x = np.arange(6)
z = x[[1, 2, 3]]
z[0] = 99
x
np.shares_memory(x, z)
```
:::
::::

::: {.aside}
Use `.copy()` when you need an independent array. Views are efficient, but mutations can propagate unexpectedly.
:::


## Exercise 1

Without running the code, determine the result (or error) of each expression.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
#| eval: false
x = c(a = 2, b = 4, c = 6, d = 8)
```
```{r}
#| eval: false
x[c(1, 3)]
x[-c(1, 3)]
x[c(TRUE, FALSE)]
x[c("d", "b", "z")]
x[x > 3 & x < 8]
```
:::

::: {.column width='50%'}
```{python}
#| eval: false
x = np.array([2, 4, 6, 8])
```
```{python}
#| eval: false
x[[0, 2]]
x[-2:]
x[np.array([True, False])]
x[x > 3 & x < 8]
x[(x > 3) & (x < 8)]
```
:::
::::

```{r}
#| echo: false
countdown::countdown(minutes = 4)
```


# Matrices & arrays

## R matrices and NumPy arrays

Both are homogeneous, multidimensional containers designed for vectorized computation.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = matrix(1:6, nrow = 2, ncol = 3)
x
typeof(x)
dim(x)
```
:::

::: {.column width='50%'}
```{python}
x = np.array([[1, 2, 3],
              [4, 5, 6]])
x
x.dtype
x.shape
```
:::
::::

::: {.aside}
An R matrix is an atomic vector with a `dim` attribute. A NumPy array is an `ndarray` object with a fixed shape and a `dtype`.
:::


## Creating arrays

:::: {.columns .mxsmall}
::: {.column width='50%'}
```{r}
matrix(0, nrow = 2, ncol = 3)
array(1, dim = c(2, 2, 2))
diag(3)
seq(0, 1, length.out = 5)
```
:::

::: {.column width='50%'}
```{python}
np.zeros((2, 3))
np.ones((2, 2, 2))
np.eye(3)
np.linspace(0, 1, 5)
```
:::
::::



## Column-major vs row-major

R matrices use *column-major* ordering - a sequence of values fills the matrix down each column. NumPy arrays default to *row-major* ordering - values fill across each row.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = matrix(1:6, nrow = 2)
x
```
:::

::: {.column width='50%'}
```{python}
x = np.arange(1, 7).reshape((2, 3))
x
```
:::
::::

. . .

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
y = matrix(1:6, nrow = 3, byrow = TRUE)
y
c(y)
```
:::

::: {.column width='50%'}
```{python}
y = np.array([[1, 2],
              [3, 4],
              [5, 6]])
y
y.flatten()
```
:::
::::

::: {.aside}
`byrow = TRUE` changes how the matrix is filled, not how it is stored
:::


## Two-dimensional indexing

Dimensions are separated by commas in both languages; ranges select rectangular regions.

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x = matrix(1:16, nrow = 4, byrow = TRUE); x
```
:::

::: {.column width='50%'}
```{python}
x = np.arange(1, 17).reshape(4, 4); x
```
:::
::::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{r}
x[1, 2]
x[2:3, 2:4]
x[c(1, 3), c(2, 4)]
```
:::

::: {.column width='50%'}
```{python}
x[0, 1]
x[1:3, 1:4]
x[np.ix_([0, 2], [1, 3])]
```
:::
::::



## Integer arrays in 2D

An integer list can select along either dimension. When both dimensions receive integer lists, NumPy pairs them element by element instead of crossing them.


:::: {.columns .small}
::: {.column width='50%'}
```{python}
x = np.arange(1, 17).reshape(4, 4)
x[[0, 2], :]
x[:, [1, 3]]
```
:::

::: {.column width='50%'}
```{python}
x[[0, 2], [1, 3]]
x[np.ix_([0, 2], [1, 3])]
```
:::
::::

. . .

The paired expression selects `(row 0, column 1)` and `(row 2, column 3)`, not a $2 \times 2$ rectangle; `np.ix_()` produces the rectangle.


## Reductions and axes

A *reduction* combines many values into fewer values. NumPy's `axis` argument specifies which dimension is collapsed.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = matrix(1:12, nrow = 3, byrow = TRUE); x
```
:::

::: {.column width='50%'}
```{python}
x = np.arange(1, 13).reshape(3, 4); x
```
:::
::::

:::: {.columns .small}
::: {.column width='50%'}
```{r}
sum(x)
rowMeans(x)
colMeans(x)
```
:::

::: {.column width='50%'}
```{python}
x.sum()
x.mean(axis=1)
x.mean(axis=0)
```
:::
::::

::: {.aside}
In NumPy, `axis=0` collapses rows; `axis=1` collapses columns.
:::


# NumPy broadcasting


## NumPy broadcasts by shape

R recycles by comparing lengths; NumPy *broadcasts* by comparing shapes. 

Shapes are compared from the rightmost dimension to the left, and each pair of dimensions is compatible when they are equal or one is `1`.

::: {.xsmall}
```{python}
x = np.arange(1, 7)
```
:::

:::: {.columns .xsmall}
::: {.column width='50%'}
```{python}
x + 10
x + np.array([10])
```
:::

::: {.column width='50%'}
```{python}
#| error: true
x + np.array([10, 100])
```
```{python}
x + np.array([10, 100, 1000,
              10_000, 100_000, 1_000_000])
```
:::
::::

::: {.aside}
Compare with R: `1:6 + c(10, 100)` recycles silently, but the shapes `(6,)` and `(2,)` are incompatible in NumPy.
:::


## Broadcasting across rows

A one-dimensional array aligns with the trailing dimension, so it operates across every row.

::: {.small}
```{python}
x = np.arange(12).reshape(4, 3)
y = np.array([10, 20, 30])
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
x
x.shape
y
y.shape
```
:::

::: {.column .fragment width='50%'}
```{python}
x + y
(x + y).shape
```
:::
::::


## Broadcasting down columns

To align with rows, give the vector an explicit singleton column dimension.

::: {.small}
```{python}
x = np.arange(12).reshape(3, 4)
y = np.array([10, 20, 30])
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
y.shape
y[:, np.newaxis]
y[:, np.newaxis].shape
```
:::

::: {.column width='50%'}
```{python}
x + y[:, np.newaxis]
```
:::
::::

::: {.aside}
`np.newaxis` is just an alias for `None`, so the equivalent `y[:, None]` is common in real-world code.
:::

## Pairwise combinations

Broadcasting two singleton dimensions produces every pairwise combination.

:::: {.columns .small}
::: {.column width='50%'}
```{r}
x = 1:3
y = c(10, 20, 30, 40)
outer(x, y, FUN = "+")
```
:::

::: {.column width='50%'}
```{python}
x = np.arange(1, 4)
y = np.arange(10, 50, 10)
x[:, np.newaxis] + y
```
:::
::::

::: {.aside}
The NumPy operands have shapes `(3, 1)` and `(4,)`, so the result has shape `(3, 4)`.
:::


## Standardizing columns

Broadcasting makes it possible to transform every column using its own mean and standard deviation.

::: {.small}
```{python}
rng = np.random.default_rng(523)
x = rng.normal(
    loc=[-1, 0, 1],
    scale=[1, 2, 3],
    size=(1000, 3)
)
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
means = x.mean(axis=0)
sds = x.std(axis=0)
means
sds
```
:::

::: {.column width='50%'}
```{python}
z = (x - means) / sds
z.mean(axis=0)
z.std(axis=0)
```
:::
::::


## Unintended broadcasting

Valid shapes do not guarantee the intended calculation.

::: {.small}
```{python}
x = np.array([[1, 2, 3],
              [4, 5, 6],
              [7, 8, 9]])
row_means = x.mean(axis=1)
```
:::

:::: {.columns .small}
::: {.column width='50%'}
```{python}
x - row_means
```
:::

::: {.column width='50%'}
```{python}
x - row_means[:, np.newaxis]
```
:::
::::

. . .

Both run, but only the second subtracts each row's own mean.

::: {.aside}
When an expression is surprising, inspect `.shape` before the values. `np.broadcast_shapes()` can predict the result shape without performing the calculation.
:::


## Exercise 2

For each pair of NumPy shapes, determine whether broadcasting succeeds and, if so, the result shape.

::: {.medium}
1. `(128, 128, 3)` and `(3,)`

2. `(8, 1, 6, 1)` and `(7, 1, 5)`

3. `(2, 1)` and `(8, 4, 3)`

4. `(3, 1)` and `(15, 3, 5)`

5. `(3,)` and `(4,)`
:::

Then write a vectorized NumPy expression that subtracts the mean of each row from a matrix `x` with shape `(100, 5)`.

```{r}
#| echo: false
countdown::countdown(minutes = 5)
```


# Comparing R & Python

## Vectorization summary

::: {.small}
| Concept                  | R                                  | Python / NumPy                        |
|:-------------------------|:-----------------------------------|:--------------------------------------|
| element-wise arithmetic  | built into atomic vectors          | NumPy arrays                          |
| element-wise comparisons | built into atomic vectors          | NumPy arrays                          |
| logical operators        | `&`, <code>&#124;</code>, `!`       | `&`, <code>&#124;</code>, `~`         |
| element-wise choice      | `ifelse()`                         | `np.where()`                          |
| scalar repetition        | recycling                          | broadcasting                          |
| unequal sizes            | repeat by total length             | compare shapes from the right         |
| explicit iteration       | `for`, later `lapply()` / `purrr`  | comprehension, `for`                  |
| reduction                | `sum()`, `mean()`, `rowMeans()`    | `.sum()`, `.mean(axis=...)`           |
:::


## Subsetting summary

::: {.small}
| Goal                    | R                                | Python / NumPy                         |
|:------------------------|:---------------------------------|:---------------------------------------|
| first element           | `x[1]`                           | `x[0]`                                 |
| last element            | `x[length(x)]`                   | `x[-1]`                                |
| exclude first           | `x[-1]`                          | `x[1:]`                                |
| regular range           | `x[2:5]`                         | `x[1:5]`                               |
| arbitrary positions     | `x[c(1, 3)]`                     | `x[[0, 2]]` (NumPy)                    |
| Boolean filter          | `x[x > 0]`                       | `x[x > 0]` (NumPy)                     |
| all rows, second column | `x[, 2]`                         | `x[:, 1]`                              |
| independent copy        | usually automatic on modification | `x.copy()` (NumPy)                    |
:::


## Takeaways

* R atomic vectors and NumPy arrays support element-wise operations; Python lists require explicit iteration or comprehensions.

* R indexes from 1 and uses negative indexes for exclusion; Python indexes from 0, counts backward with negative indexes, and uses half-open slices.

* R's `[` supports integer, logical, and name-based selection. NumPy adds integer-array and Boolean indexing to Python's usual indexing syntax.

* Basic NumPy slices are usually views; advanced indexing is a copy. Make copying intentional when the result will be modified.

* R recycling is based on vector length; NumPy broadcasting is based on compatible shapes. When in doubt, inspect `.shape`.
