---
title: "Lecture 09 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
```

```{python py_setup}
#| include: false
import polars as pl
```


# Exercise 1

Using polars and `flights.parquet`, repeat the questions from last lecture's pandas exercise.

```{python}
flights = pl.read_parquet("../data/flights.parquet")
```

1. How many flights to LAX did each legacy carrier (AA, UA, DL, US) have in May from JFK, and what was their mean actual air time?

   ```{python}
   flights.filter(
     pl.col("dest") == "LAX",
     pl.col("origin") == "JFK",
     pl.col("month") == 5,
     pl.col("carrier").is_in(["AA", "UA", "DL", "US"])
   ).group_by("carrier").agg(
     n = pl.len(),
     avg_air_time = pl.col("air_time").mean()
   ).sort("carrier")
   ```

2. Which plane has the most flight records from each New York airport (no missing tail numbers, all ties)?

   ```{python}
   (flights
     .drop_nulls("tailnum")
     .group_by("origin", "tailnum")
     .len()
     .filter(pl.col("len") == pl.col("len").max().over("origin"))
     .sort("origin")
   )
   ```

3. Which five calendar dates had the lowest mean departure delay (ties broken by earliest date)?

   ```{python}
   (flights
     .group_by("year", "month", "day")
     .agg(avg_dep_delay = pl.col("dep_delay").mean())
     .drop_nulls("avg_dep_delay")
     .sort("avg_dep_delay", "year", "month", "day")
     .head(5)
   )
   ```

4. Which flight has the largest arrival delay as a percentage of its actual air time (no missing values or nonpositive air times, all ties)?

   ```{python}
   (flights
     .filter(
      pl.col("arr_delay").is_not_null(), pl.col("air_time") > 0
    )
     .with_columns(
      arr_delay_pct = 100 * pl.col("arr_delay") / pl.col("air_time")
    )
     .filter(
      pl.col("arr_delay_pct") == pl.col("arr_delay_pct").max()
    )
     .select(
      "month", "day", "carrier", "flight", "origin", "dest", "arr_delay", "air_time", "arr_delay_pct"
    )
   )
   ```


# Exercise 2

Using tidyr and then polars, construct a contingency table of penguin counts with islands as rows and species as columns, with zeros for the missing pairs.

```{r}
palmerpenguins::penguins |>
  count(island, species) |>
  pivot_wider(
    names_from = species,
    values_from = n,
    values_fill = 0
  )
```

```{python}
(pl.read_csv("../data/penguins.csv", null_values="NA")
  .group_by("island", "species")
  .len()
  .pivot(on="species", index="island", values="len")
  .fill_null(0)
  .sort("island")
)
```


# Exercise 3

Using dplyr and then polars, with `flights.parquet` and `planes.parquet`:

```{r}
flights = nanoparquet::read_parquet("../data/flights.parquet")
planes = nanoparquet::read_parquet("../data/planes.parquet")
```

```{python}
planes = pl.read_parquet("../data/planes.parquet")
```

1. How many flights have a tail number with no matching record in `planes`? Which carrier accounts for the most of them?

   ```{r}
   unmatched = flights |>
     filter(!is.na(tailnum)) |>
     anti_join(planes, by = "tailnum")

   nrow(unmatched)
   count(unmatched, carrier, sort = TRUE)
   ```

   ```{python}
   unmatched = (flights
     .drop_nulls("tailnum")
     .join(planes, on="tailnum", how="anti")
   )

   unmatched.height
   unmatched.group_by("carrier").len().sort("len", descending=True)
   ```

2. Which five manufacturers' planes flew the most flights out of New York?

   ```{r}
   flights |>
     inner_join(planes, by = "tailnum") |>
     count(manufacturer, sort = TRUE) |>
     head(5)
   ```

   ```{python}
   (flights
     .join(planes, on="tailnum")
     .group_by("manufacturer")
     .len()
     .sort("len", descending=True)
     .head(5)
   )
   ```

3. Both tables have a `year` column. What happens to these columns in a join, and do they mean the same thing?

   ```{r}
   flights |>
     left_join(planes, by = "tailnum") |>
     select(tailnum, starts_with("year"))
   ```

   ```{python}
   (flights
     .join(planes, on="tailnum", how="left")
     .select("tailnum", pl.selectors.starts_with("year"))
   )
   ```

   Non-key columns with the same name get a suffix (`year.x` / `year.y`, `year` / `year_right`). `flights$year` is the year of the flight, `planes$year` is the year of manufacture, so they should not be used as a key.
