---
title: "Lecture 06 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```


# Exercise 1

Represent the JSON record in Python, extract the fax number, add a mobile number and remove `age`, then write a `phone()` lookup function in both languages.

## Building the structure

JSON objects become `dict`s and JSON arrays become `list`s. Since JSON is valid Python syntax (for this record at least - `true` / `false` / `null` would need to become `True` / `False` / `None`) the text can be pasted in directly.

```{python}
person = {
  "firstName": "John",
  "lastName": "Smith",
  "age": 25,
  "address": {
    "streetAddress": "21 2nd Street",
    "city": "New York",
    "state": "NY",
    "postalCode": 10021
  },
  "phoneNumber": [
    {"type": "home", "number": "212 555-1239"},
    {"type": "fax",  "number": "646 555-4567"}
  ]
}
```

In practice the `json` module does this conversion - `json.loads(text)` returns exactly this structure and `json.dumps(person)` goes back the other way. R's equivalent is `jsonlite::fromJSON(text, simplifyVector = FALSE)`.

## Extracting the fax number

`phoneNumber` is a list, so its entries are selected by position (0-based), and each entry is a `dict` selected by key,

```{python}
person["phoneNumber"][1]["number"]
```

Compare with the R version, `person$phoneNumber[[2]]$number` - the same shape, with `[]` doing the work of both `[[` and `$`.

## Adding a mobile number and removing `age`

```{python}
person["phoneNumber"].append({"type": "mobile", "number": "917 555-0000"})
del person["age"]

list(person)
len(person["phoneNumber"])
```

Both operations modify `person` in place, where the R versions (`person$phoneNumber[[3]] = ...` and `person$age = NULL`) modify a copy that is then rebound to the name `person`. Note that `del person["age"]` would raise a `KeyError` if run a second time, `person.pop("age", None)` is the safe alternative.

## `phone()` in Python

The phone numbers are a list of dictionaries, so finding one requires a scan - return as soon as a match is found, and fall through to `None` if there is none.

```{python}
def phone(person, type):
    """Return the phone number of the given type, or None if there is not one."""
    for entry in person["phoneNumber"]:
        if entry["type"] == type:
            return entry["number"]
    return None
```

```{python}
phone(person, "fax")
print( phone(person, "work") )
```

If lookups by type were common the better design is to store the numbers as a `dict` keyed by type in the first place - then `numbers.get(type)` does the job with no loop,

```{python}
numbers = {entry["type"]: entry["number"] for entry in person["phoneNumber"]}
numbers
numbers.get("fax")
```

## `phone()` in R

The direct translation loops over the list and uses `return()`, falling through to `NULL`,

```{r}
person = list(
  firstName = "John", lastName = "Smith",
  address = list(streetAddress = "21 2nd Street", city = "New York", state = "NY", postalCode = 10021),
  phoneNumber = list(
    list(type = "home",   number = "212 555-1239"),
    list(type = "fax",    number = "646 555-4567"),
    list(type = "mobile", number = "917 555-0000")
  )
)

phone = function(person, type) {
  for (entry in person$phoneNumber) {
    if (entry$type == type)
      return(entry$number)
  }
  NULL
}
```

```{r}
phone(person, "fax")
phone(person, "work")
```

A more R-like approach extracts the vector of types and uses `match()`, which is what we would do if there were many lookups to perform,

```{r}
types = vapply(person$phoneNumber, function(entry) entry$type, character(1))
types
match("fax", types)
person$phoneNumber[[ match("fax", types) ]]$number
```

Note that `match()` returns `NA` for a missing type, and `person$phoneNumber[[NA]]` is an error, so this version needs an explicit check before subsetting. Next week's functional programming tools (`purrr::map_chr()`, `purrr::keep()`) make this pattern much cleaner.


# Exercise 2

For each scenario, which data structure is the most appropriate and why?

## A fixed collection of 100 integers

Python: a `tuple` - the collection does not change, so immutability documents that intent and lets it be used as a `dict` key. If arithmetic is needed, a NumPy array.

R: an integer atomic vector (`integer(100)`, or `1:100`). R has no immutable containers, but copy on modify means that nothing else can change it out from under you.

## A queue (first in, first out) of customer records

Python: a `deque`. `append()` to add at the back and `popleft()` to serve from the front are both O(1), while `list.pop(0)` has to shift every remaining element.

R: there is no efficient built in option. A `list` with `x[[length(x) + 1]] = record` to add and `x = x[-1]` to remove works but the removal copies the list (O(n)). For a large queue keep an index of the current head instead of removing elements, or use an environment based implementation from a package.

## A stack (first in, last out) of customer records

Python: a plain `list` - `append()` and `pop()` at the end are O(1).

R: a `list`, adding with `x[[length(x) + 1]] = record` and removing with `x[[length(x)]]` / `x = x[-length(x)]`. Removing the last element still copies, but this is the best base R can do.

## A count of word occurrences within a document

Python: a `dict` mapping word to count (or `collections.Counter`, which is a `dict` subclass). Each update is an O(1) lookup.

R: `table(words)` for a one-off count. For incremental counting a named integer vector works for small vocabularies, an environment (`new.env()`) is the hash-based option for large ones.

## The heights of the bars in a histogram with even bin widths

Python: a `list` (or NumPy array) of counts - the position in the sequence identifies the bin, so no keys are needed.

R: a numeric vector, for the same reason. In both cases a `dict` / named list would be wasteful since the keys are just `0, 1, 2, ...`.

## Checking whether each of a million words is in a list of 50 stop words

Python: put the stop words in a `set` and test `word in stop_words` - O(1) per word rather than O(50) with a `list`. The million words themselves can stay in a `list`.

R: `words %in% stop_words`. `%in%` (via `match()`) builds a hash table of its second argument internally, so this is already the efficient approach - and it is vectorized, so no loop is needed.

## The last 10 lines read from a very large log file

Python: `deque(maxlen=10)` - appending a line when full automatically drops the oldest, so the whole file never needs to be held in memory.

R: there is no `maxlen` equivalent. Keep a character vector of length 10 and shift with `x = c(x[-1], line)` (cheap for a length 10 vector), or if the file fits in memory simply `tail(readLines(path), 10)`.
