---
title: "Lecture 03 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```


# Exercise 1

Without running the code, what do you expect the output (or error) to be for each of the listed values of `x`?

To make it easier to try each value, the conditional block from the slides is wrapped in a function below - the behavior is identical to assigning `x` and running the block at the top level.

## R

```{r}
check = function(x) {
  if (x > 10 || x < -10) {
    stop("Input too big")
  } else if (x %in% c(2, 3, 5, 7)) {
    cat("Input is prime!\n")
  } else if (x %% 2 == 0) {
    cat("Input is even!\n")
  } else if (x %% 2 == 1) {
    cat("Input is odd!\n")
  }
}
```

```{r}
check(1)       # 1 %% 2 == 1                          -> odd
check(3)       # 3 %in% c(2, 3, 5, 7)                 -> prime
check(-1)      # -1 %% 2 == 1 (%% takes the sign of the divisor) -> odd
check(-3)      # -3 %% 2 == 1                         -> odd
```

```{r}
#| error: true
check(1:2)     # x > 10 has length 2, || requires length 1 (R >= 4.3) -> error
```

```{r}
#| error: true
check("0")     # "0" > 10 and "0" < -10 are string comparisons ("0" > "10", "0" < "-10"),
               # both FALSE; "0" %in% c(2, 3, 5, 7) is FALSE (compared as strings);
               # "0" %% 2 does not coerce character to numeric   -> error
```

```{r}
#| error: true
check("zero")  # "zero" > 10 compares as strings, "zero" > "10" is TRUE
               # (letters sort after digits) so || short-circuits -> stop()
```

Note that for `"0"` and `"zero"` the code fails, but at different points and for different reasons - string comparison silently "works" in R (via coercion of the number to a string) while arithmetic does not.


## Python

```{python}
def check(x):
    if x > 10 or x < -10:
        raise ValueError("Input too big")
    elif x in [2, 3, 5, 7]:
        print("Input is prime!")
    elif x % 2 == 0:
        print("Input is even!")
    elif x % 2 == 1:
        print("Input is odd!")
```

```{python}
check(1)       # 1 % 2 == 1                           -> odd
check(3)       # 3 in [2, 3, 5, 7]                    -> prime
check(-1)      # -1 % 2 == 1 (% takes the sign of the divisor) -> odd
check(-3)      # -3 % 2 == 1                          -> odd
check(2.5)     # 2.5 % 2 == 0.5, no branch matches    -> no output
```

```{python}
#| error: true
check([1, 2])  # list > int is not supported -> TypeError
```

```{python}
#| error: true
check("0")     # str > int is not supported (no implicit coercion) -> TypeError
```

Unlike R, Python refuses to compare a string (or list) with a number, so both of the "bad" inputs fail immediately at the first condition with the same error type.


# Exercise 2

Using *nested* loops, write code in both R and Python that prints only the values of `x` that are *not* prime - without using subsetting, `%in%`, or `in`.

```{r}
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 
           29, 31, 37, 41, 43, 47, 53, 59, 61, 
           67, 71, 73, 79, 83, 89, 97)
x = c(3, 4, 12, 19, 23, 51, 61, 63, 78)
```

```{python}
primes = [ 2,  3,  5,  7, 11, 13, 17, 19, 23, 
          29, 31, 37, 41, 43, 47, 53, 59, 61, 
          67, 71, 73, 79, 83, 89, 97]
x = [3, 4, 12, 19, 23, 51, 61, 63, 78]
```

## R

R has no loop `else`, so we need a flag variable that is set if a match is found in the inner loop. The `break` is not strictly necessary but avoids checking the remaining primes once a match is found.

```{r}
for (v in x) {
  is_prime = FALSE
  
  for (p in primes) {
    if (v == p) {
      is_prime = TRUE
      break
    }
  }
  
  if (!is_prime)
    print(v)
}
```

## Python

The same flag-based approach works in Python,

```{python}
for v in x:
    is_prime = False

    for p in primes:
        if v == p:
            is_prime = True
            break

    if not is_prime:
        print(v)
```

but the loop `else` clause removes the need for the flag - the `else` body runs only when the inner loop finishes without hitting `break`, i.e. when no prime matched.

```{python}
for v in x:
    for p in primes:
        if v == p:
            break
    else:
        print(v)
```

## Without loops

For comparison, the idiomatic (non-loop) solutions in each language, using the operators the exercise asked you to avoid:

```{r}
x[!x %in% primes]
```

```{python}
[v for v in x if v not in primes]
```
