---
title: "Lecture 05 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```


# Exercise 1

Represent the JSON record as a list in R, then extract the fax number, add a mobile number, and remove the `age` element.

## Building the list

JSON objects (`{}`) become named lists and JSON arrays (`[]`) become unnamed lists. Atomic values stay as length one vectors.

```{r}
person = list(
  firstName = "John",
  lastName = "Smith",
  age = 25,
  address = list(
    streetAddress = "21 2nd Street",
    city = "New York",
    state = "NY",
    postalCode = 10021
  ),
  phoneNumber = list(
    list(type = "home", number = "212 555-1239"),
    list(type = "fax",  number = "646 555-4567")
  )
)

str(person)
```

The `phoneNumber` array could also be represented as a data frame (two columns, two rows) since every entry has the same fields - this is what `jsonlite::fromJSON()` does by default. The list of lists above is the more literal translation and is what `jsonlite::fromJSON(txt, simplifyVector = FALSE)` returns.

## Extracting the fax number

`phoneNumber` is an unnamed list, so its elements must be selected by position with `[[`. The elements inside are named lists, so `$` or `[[` with a name both work.

```{r}
person$phoneNumber[[2]]$number
person[["phoneNumber"]][[2]][["number"]]
```

Recursive indexing with a vector of positions also works (`phoneNumber` is the 5th element, the fax entry is its 2nd, and `number` is the 2nd element of that), but it is much harder to read and names cannot be mixed with positions,

```{r}
person[[c(5, 2, 2)]]
```

## Adding a mobile number

Assigning to a position one past the end of a list appends an element,

```{r}
person$phoneNumber[[3]] = list(type = "mobile", number = "917 555-0000")
str(person$phoneNumber)
```

`append()` is an alternative, but note that the new element must itself be wrapped in a `list()` since `append()` concatenates its arguments,

```{r}
str( append(person$phoneNumber, list(list(type = "work", number = "212 555-0000"))) )
```

## Removing `age`

Assigning `NULL` removes an element,

```{r}
person$age = NULL
names(person)
```

To store an actual `NULL` value instead use `person["age"] = list(NULL)`.


# Exercise 2

Implement a `circle` class in Python and the equivalent S3 class in R.

## Python

```{python}
import math

class circle:
    """A circle with radius r"""

    def __init__(self, r=1):
        if r < 0:
            raise ValueError("r must be non-negative")
        self.r = r

    def area(self):
        return math.pi * self.r**2

    def __repr__(self):
        return f"circle(r={self.r})"
```

```{python}
circle(2)
circle(2).area()
[circle(), circle(0.5)]
```

```{python}
#| error: true
circle(-1)
```

Note that because `__str__()` is not defined, `print(circle(2))` falls back to `__repr__()` and shows the same thing. The validation could be extended with `isinstance(r, (int, float))` to reject non-numeric inputs, keeping in mind that this would also accept `True` and `False`.

## R

The constructor validates its input and attaches the class to a list. The `area()` generic was defined on the slides - it is repeated here so this document is self contained.

```{r}
circle = function(r = 1) {
  stopifnot(is.numeric(r), length(r) == 1)
  if (r < 0)
    stop("`r` must be non-negative.", call. = FALSE)

  structure(list(r = r), class = "circle")
}

area = function(x, ...) UseMethod("area")
area.circle = function(x, ...) pi * x$r^2

print.circle = function(x, ...) {
  cat("circle(r = ", x$r, ")\n", sep = "")
  invisible(x)
}
```

```{r}
circle(2)
area(circle(2))
str(unclass(circle(2)))
```

```{r}
#| error: true
circle(-1)
area(2)
```

The last error comes from `UseMethod()` - there is no `area.numeric()` or `area.default()` method, so dispatch fails. Adding an `area.default()` that calls `stop()` with a friendlier message is good practice.

Finally, a check that the print method returns its input invisibly,

```{r}
x = print(circle(3))
x$r
```
