---
title: "Lecture 02 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```


# Exercise 1

What is the type of each of the R vectors and why, and what is the value (and type) of each of the Python expressions? How does R's implicit conversion hierarchy for `logical`, `integer`, `double`, and `character` compare to Python's numeric promotion rules?

```{r}
typeof(c(1, NA+1L, "C"))   # NA + 1L is NA_integer_, character is the highest type present
typeof(c(1L / 0, NA))      # 1L / 0 is Inf, a double
typeof(c(1:3, 5))          # 5 is a double literal
typeof(c(3L, NaN+1L))      # NaN + 1L is NaN, a double
typeof(c(NA, TRUE))        # NA is logical by default
typeof(c(1L, NULL, NA))    # NULL contributes nothing, NA is coerced to NA_integer_
```

```{python}
True + 1.5                # bool promotes to float
"1" * 3                   # str * int repeats the string
int("3") + float("1.5")   # int + float promotes to float
bool("False")             # any non-empty string is truthy
int(True) / True          # / always returns a float
```

R: `logical` < `integer` < `double` < `character`, a mixed vector is coerced to the highest type present.

Python: `bool` < `int` < `float` < `complex`, promotion only happens between numeric types and mixing a number with a `str` raises a `TypeError` rather than coercing.


# Exercise 2

Without running the code, predict the output (or error) of each of the following in both R and Python.

## R

```{r}
x = c(1, 2, 3)
y = x
y[1] = "a"               # y is copied on modify and coerced to character, x is unchanged
x
y
```

```{r}
TRUE + TRUE == 2         # TRUE is coerced to 1
"2" == 2                 # 2 is coerced to "2"
as.integer("2") + 1L
0.1 + 0.2 == 0.3         # 0.1 + 0.2 is 0.30000000000000004
```

```{r}
c(1, 2.5, NA)            # NA is coerced to NA_real_
sum(c(1, 2.5, NA))       # NA is sticky
```

## Python

```{python}
x = [1, 2, 3]
y = x
y[0] = "a"               # y is another name for the same list, so x changes too
x
y
```

```{python}
True + True == 2         # bool is a subclass of int
"2" == 2                 # no coercion, values of different types are not equal
int("2") + 1
0.1 + 0.2 == 0.3         # 0.1 + 0.2 is 0.30000000000000004
```

```{python}
[1, 2.5, None]           # lists are heterogeneous, nothing is coerced
```

```{python}
#| error: true
sum([1, 2.5, None])      # float + None is not defined
```
