---
title: "Lecture 05 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```


# Exercise 1

Represent the JSON record as a list in R, then extract the fax number, add a mobile number, and remove the `age` element.

## Building the list

JSON objects (`{}`) become named lists and JSON arrays (`[]`) become unnamed lists. Atomic values stay as length one vectors.

```{r}
person = list(
  firstName = "John",
  lastName = "Smith",
  age = 25,
  address = list(
    streetAddress = "21 2nd Street",
    city = "New York",
    state = "NY",
    postalCode = 10021
  ),
  phoneNumber = list(
    list(type = "home", number = "212 555-1239"),
    list(type = "fax",  number = "646 555-4567")
  )
)

str(person)
```

The `phoneNumber` array could also be represented as a data frame (two columns, two rows) since every entry has the same fields - this is what `jsonlite::fromJSON()` does by default. The list of lists above is the more literal translation and is what `jsonlite::fromJSON(txt, simplifyVector = FALSE)` returns.

## Extracting the fax number

`phoneNumber` is an unnamed list, so its elements must be selected by position with `[[`. The elements inside are named lists, so `$` or `[[` with a name both work.

```{r}
person$phoneNumber[[2]]$number
person[["phoneNumber"]][[2]][["number"]]
```

Recursive indexing with a vector of positions also works (`phoneNumber` is the 5th element, the fax entry is its 2nd, and `number` is the 2nd element of that), but it is much harder to read and names cannot be mixed with positions,

```{r}
person[[c(5, 2, 2)]]
```

## Adding a mobile number

Assigning to a position one past the end of a list appends an element,

```{r}
person$phoneNumber[[3]] = list(type = "mobile", number = "917 555-0000")
str(person$phoneNumber)
```

`append()` is an alternative, but note that the new element must itself be wrapped in a `list()` since `append()` concatenates its arguments,

```{r}
str( append(person$phoneNumber, list(list(type = "work", number = "212 555-0000"))) )
```

## Removing `age`

Assigning `NULL` removes an element,

```{r}
person$age = NULL
names(person)
```

To store an actual `NULL` value instead use `person["age"] = list(NULL)`.


