---
title: "Lecture 04 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```

```{python py_setup}
#| echo: false
import numpy as np
```


# Exercise 1

Without running the code, determine the result (or error) of each expression.

## R

```{r}
x = c(a = 2, b = 4, c = 6, d = 8)
```

Positive positions select the first and third elements and preserve their names.

```{r}
x[c(1, 3)]
```

Negative positions exclude the first and third elements.

```{r}
x[-c(1, 3)]
```

The two-value logical index is recycled to the length of `x`, so it keeps positions 1 and 3.

```{r}
x[c(TRUE, FALSE)]
```

Character indexing looks up names. The unknown name `"z"` produces `NA` and retains the requested name.

```{r}
x[c("d", "b", "z")]
```

The comparisons and `&` are vectorized, so this keeps values strictly between 3 and 8.

```{r}
x[x > 3 & x < 8]
```


## Python / NumPy

```{python}
x = np.array([2, 4, 6, 8])
```

An integer list triggers NumPy advanced indexing and selects positions 0 and 2.

```{python}
x[[0, 2]]
```

The slice begins two positions from the end and continues through the end.

```{python}
x[-2:]
```

A Boolean index must match the dimension being indexed, so this raises `IndexError` rather than recycling.

```{python}
#| error: true
x[np.array([True, False])]
```

Without parentheses, Python's operator precedence parses the chained comparison in an unintended way and NumPy raises an error about the truth value of an array.

```{python}
#| error: true
x[x > 3 & x < 8]
```

Parenthesizing each comparison produces the intended element-wise Boolean mask.

```{python}
x[(x > 3) & (x < 8)]
```


# Exercise 2

NumPy compares shapes from right to left. At each position, dimensions are compatible if they are equal or one of them is 1.

## Shape 1

```text
  (128, 128, 3)
            (3,)
  -------------
  (128, 128, 3)
```

This succeeds. The `(3,)` array broadcasts over the two leading dimensions.


## Shape 2

```text
  (8, 1, 6, 1)
     (7, 1, 5)
  -------------
  (8, 7, 6, 5)
```

This succeeds: comparing from the right gives `1` with `5`, `6` with `1`, `1` with `7`, and the missing leading dimension with `8`.


## Shape 3

```text
     (2, 1)
  (8, 4, 3)
  ---------
  incompatible
```

This fails because the trailing dimensions `1` and `3` are compatible, but `2` and `4` are not.


## Shape 4

```text
     (3, 1)
  (15, 3, 5)
  ----------
  (15, 3, 5)
```

This succeeds: `1` expands to `5`, `3` matches `3`, and the missing leading dimension expands to `15`.


## Shape 5

```text
  (3,)
  (4,)
  ----
  incompatible
```

This fails because `3` and `4` are unequal and neither is 1.


## Centering each row

For an array `x` with shape `(100, 5)`, `x.mean(axis=1)` has shape `(100,)`. That cannot align with the trailing dimension of size 5, so add a singleton column dimension before subtracting.

```{python}
rng = np.random.default_rng(523)
x = rng.normal(size=(100, 5))

centered = x - x.mean(axis=1)[:, np.newaxis]
centered.shape
centered.mean(axis=1)[:10]
```

Equivalently, keep the collapsed dimension during the reduction:

```{python}
centered = x - x.mean(axis=1, keepdims=True)
centered.shape
```
