---
title: "Lecture 02 - Exercise solutions"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: true
engine: knitr
---

```{r setup}
#| include: false
options(width = 80)
```


# Exercise 1

## Part 1

What is the type of the following R vectors? Explain why they have that type.

```{r}
c(1, NA+1L, "C")   # c(1, NA_integer_, "C") -> character
c(1L / 0, NA)      # c(Inf, NA)              -> double
c(1:3, 5)          # c(1L, 2L, 3L, 5)        -> double
c(3L, NaN+1L)      # c(3L, NaN)              -> double
c(NA, TRUE)        # c(NA, TRUE)             -> logical
c(1L, NULL, NA)    # c(1L, NA)               -> integer
```


## Part 2

Without running the code, what is the value (and type) of each of the following Python expressions?

```{python}
True + 1.5                # 2.5   -> float
"1" * 3                   # '111' -> str
int("3") + float("1.5")   # 4.5   -> float
bool("False")             # True  -> bool
int(True) / True          # 1.0   -> float
```


## Part 3

Considering R's four most commonly used atomic types (`logical`, `integer`, `double`, and `character`), what is the implicit conversion hierarchy from highest to lowest priority? How does this compare to Python's numeric promotion rules?

R:
`character` > `double` > `integer` > `logical`

Python:
`complex` > `float` > `int` > `bool`


# Exercise 2

Without running the code, predict the output (or error) of each of the following in *both* R and Python.

## Assignment and coercion

### R

```{r}
x = c(1, 2, 3)
y = x                    # copy on modify
y[1] = "a"               # coerces y to character
x                        # c(1, 2, 3)
y                        # c("a", "2", "3")
```

### Python

```{python}
x = [1, 2, 3]
y = x                    # same object
y[0] = "a"               # modifies x and y
x                        # ['a', 2, 3]
y                        # ['a', 2, 3]
```


## Comparisons and arithmetic

### R

```{r}
TRUE + TRUE == 2         # 2 == 2     -> TRUE
"2" == 2                 # "2" == "2" -> TRUE
as.integer("2") + 1L     # 2L + 1L    -> 3L
0.1 + 0.2 == 0.3         # 0.30000000000000004 == 0.3 -> FALSE
```

### Python

```{python}
True + True == 2         # 2 == 2     -> True
"2" == 2                 # str != int -> False
int("2") + 1             # 2 + 1      -> 3
0.1 + 0.2 == 0.3         # 0.30000000000000004 == 0.3 -> False
```


## Missing values in aggregation

### R

```{r}
c(1, 2.5, NA)            # c(1, 2.5, NA_real_) -> double
sum(c(1, 2.5, NA))       # NA
```

### Python

```{python}
[1, 2.5, None]           # [1, 2.5, None] -> list
```

```{python}
#| error: true
sum([1, 2.5, None])      # 3.5 + None -> TypeError
```

