```{r setup}
#| include: false
library(purrr)
library(dplyr)
library(tidyr)
```

```{python py_setup}
#| include: false
import json
import polars as pl
```


# Exercise 1

Using purrr in R and comprehensions in Python, build a data frame of each repository's name and primary language, find the five most starred repositories, and the repositories with `"r"` among their topics.

```{r}
repos = jsonlite::read_json("../data/tidyverse_repos.json")
```

```{python}
with open("../data/tidyverse_repos.json") as f:
    repos = json.load(f)
```

1. A data frame with each repository's `name` and primary `language`.

   ```{r}
   tibble(
     name = map_chr(repos, "name"),
     language = map_chr(repos, "language", .default = NA)
   )
   ```

   ```{python}
   pl.DataFrame({
     "name": [r["name"] for r in repos],
     "language": [r["language"] for r in repos]
   })
   ```

2. The five repositories with the most stars.

   ```{r}
   map_int(repos, "stargazers_count") |>
     set_names(map_chr(repos, "name")) |>
     sort(decreasing = TRUE) |>
     head(5)
   ```

   ```{python}
   top = sorted(repos, key=lambda r: r["stargazers_count"], reverse=True)[:5]
   {r["name"]: r["stargazers_count"] for r in top}
   ```

3. Repositories with `"r"` among their topics.

   ```{r}
   keep(repos, \(x) "r" %in% x$topics) |> map_chr("name")
   ```

   ```{python}
   [r["name"] for r in repos if "r" in r["topics"]]
   ```


# Exercise 2

Using tidyr / purrr in R and polars in Python, tidy `repos` into one row per repository with `name`, `description`, `language`, `stargazers_count`, `forks_count`, the license name, and the number of topics.

```{r}
tibble(repo = repos) |>
  hoist(
    repo, "name", "description", "language",
    "stargazers_count", "forks_count",
    license = c("license", "name")
  ) |>
  mutate(n_topics = map_int(repo, \(x) length(x$topics))) |>
  select(-repo)
```

```{python}
pl.DataFrame(repos).select(
  "name", "description", "language",
  "stargazers_count", "forks_count",
  license = pl.col("license").struct.field("name"),
  n_topics = pl.col("topics").list.len()
)
```
