```{r setup}
#| include: false
library(rvest)
```

```{python py_setup}
#| include: false
from urllib.parse import urljoin
from urllib.robotparser import RobotFileParser

import pandas as pd
import requests
from bs4 import BeautifulSoup

pd.set_option("display.width", 120)
```


# Example - Rotten Tomatoes

Scrape the movies present in the downloaded HTML of Movies at Home, sorted by popularity, into a data frame with the title, Tomatometer score, critic-status label, and absolute movie URL. Keep scores as text (for example, `"95%"`) and preserve missing values. Derive the critic-status label from the icon attributes.

For this example, assume each movie tile contains the title, score, critic icon, and link elements. A score element may be empty.

## BeautifulSoup

```{python}
base = "https://www.rottentomatoes.com"
url = base + "/browse/movies_at_home/sort:popular"
headers = {"User-Agent": "sta523-example"}

rp = RobotFileParser(base + "/robots.txt")
rp.read()
rp.can_fetch(headers["User-Agent"], url)
```

```{python}
response = requests.get(url, headers=headers, timeout=10)
response.raise_for_status()

page = BeautifulSoup(response.text, "html.parser")
tiles = page.select("div.flex-container")

rows = []
for tile in tiles:
    title = tile.select_one(
        "rt-text[data-qa='discovery-media-list-item-title']"
    )
    score = tile.select_one("rt-text[slot='criticsScore']")
    icon = tile.select_one("score-icon-critics")
    href = tile.select_one("a.js-tile-link")["href"]

    if icon.get("certified") == "true":
        label = "Certified Fresh"
    else:
        label = {"positive": "Fresh", "negative": "Rotten"}.get(
            icon.get("sentiment")
        )

    rows.append({
        "title": title.get_text(" ", strip=True),
        "score": score.get_text(" ", strip=True) or None,
        "label": label,
        "url": urljoin(response.url, href),
    })

movies = pd.DataFrame(rows)
movies
```

## rvest

```{r}
base = "https://www.rottentomatoes.com"
url = paste0(base, "/browse/movies_at_home/sort:popular")

session = polite::bow(url, user_agent = "sta523-example")
session
```

```{r}
page = polite::scrape(session)
tiles = page |> html_elements("div.flex-container")
icons = tiles |> html_element("score-icon-critics")

movies = tibble::tibble(
  title = tiles |> html_element("rt-text[data-qa='discovery-media-list-item-title']") |> html_text2(),
  score = tiles |> html_element("rt-text[slot='criticsScore']") |> html_text2() |> dplyr::na_if(""),
  label = dplyr::case_when(
    html_attr(icons, "certified") == "true" ~ "Certified Fresh",
    html_attr(icons, "sentiment") == "positive" ~ "Fresh",
    html_attr(icons, "sentiment") == "negative" ~ "Rotten"
  ),
  url = tiles |> html_element("a.js-tile-link") |> html_attr("href")
) |>
  dplyr::mutate(url = paste0(base, url))
movies
```
