
```{python py_setup}
#| include: false
from urllib.parse import urljoin
from urllib.robotparser import RobotFileParser

import pandas as pd
import requests
from bs4 import BeautifulSoup

pd.set_option("display.width", 120)
```


# Example - Rotten Tomatoes

Scrape the movies present in the downloaded HTML of Movies at Home, sorted by popularity, into a data frame with the title, Tomatometer score, critic-status label, and absolute movie URL. Store the score as a number (so `"95%"` becomes `95`) and preserve missing values. Derive the critic-status label from the icon attributes.

For this example, assume every movie tile contains the title, score, critic icon, and link elements, so the page-wide selections line up. A score element may be empty.

## BeautifulSoup

```{python}
base = "https://www.rottentomatoes.com"
url = base + "/browse/movies_at_home/sort:popular"
headers = {"User-Agent": "sta523-example"}

rp = RobotFileParser(base + "/robots.txt")
rp.read()
rp.can_fetch(headers["User-Agent"], url)
```

```{python}
response = requests.get(url, headers=headers, timeout=10)
response.raise_for_status()

page = BeautifulSoup(response.text, "html.parser")

movies = pd.DataFrame({
    "title": page.select("rt-text[data-qa='discovery-media-list-item-title']"),
    "score": page.select("rt-text[slot='criticsScore']"),
    "icon":  page.select("div.flex-container score-icon-critics"),
    "url":   page.select("a.js-tile-link"),
}).assign(
    title = pd.col("title").map(lambda t: t.get_text(" ", strip=True)),
    score = pd.col("score").map(lambda s: s.get_text(strip=True).rstrip("%") or None).astype("Int64"),
    sentiment = pd.col("icon").map(lambda i: i.get("sentiment")),
    certified = pd.col("icon").map(lambda i: i.get("certified") == "true"),
    label = pd.col("sentiment").map({"positive": "Fresh", "negative": "Rotten"})
              .where(~pd.col("certified"), "Certified Fresh"),
    url = pd.col("url").map(lambda a: urljoin(response.url, a["href"])),
).drop(columns=["icon", "sentiment", "certified"])
movies
```

## rvest

```{r}
#| message: false
library(tidyverse)
library(rvest)
```

```{r}
base = "https://www.rottentomatoes.com"
url = paste0(base, "/browse/movies_at_home/sort:popular")

session = polite::bow(url, user_agent = "sta523-example")
session
```

```{r}
page = polite::scrape(session)
icons = page |> html_elements("div.flex-container score-icon-critics")

movies = tibble(
  title = page |> html_elements("rt-text[data-qa='discovery-media-list-item-title']") |> html_text2(),
  score = page |> html_elements("rt-text[slot='criticsScore']") |> html_text2() |> stringr::str_remove("%") |> as.integer(),
  label = case_when(
    html_attr(icons, "certified") == "true" ~ "Certified Fresh",
    html_attr(icons, "sentiment") == "positive" ~ "Fresh",
    html_attr(icons, "sentiment") == "negative" ~ "Rotten"
  ),
  url = page |> html_elements("a.js-tile-link") |> html_attr("href")
) |>
  mutate(url = paste0(base, url))
  
movies
```
