```{r setup}
#| include: false
library(stringr)
```

```{python py_setup}
#| include: false
import re
```


# Exercise 1

For the vector of randomly generated names, write a regular expression that detects if the first name, the last name, either, or neither start with a vowel.

```{r}
x = c("Jeremy Cruz", "Nathaniel Le", "Jasmine Chu", "Bradley Calderon Raygoza",
      "Quinten Weller", "Katelien Kanamu-Hauanio", "Zuhriyaa al-Amen",
      "Travale York", "Alexis Ahmed", "David Alcocer", "Jairo Martinez",
      "Dwone Gallegos", "Amanda Sherwood", "Hadiyya el-Eid", "Shaimaaa al-Can",
      "Sarah Love", "Shelby Villano", "Sundus al-Hashmi", "Dyani Loving",
      "Shanelle Douglas")
```

```{python}
x = r.x
```

1. First name starts with a vowel.

   ```{r}
   str_subset(x, "^[AEIOU]")
   ```

   ```{python}
   [s for s in x if re.search(r"^[AEIOU]", s)]
   ```

2. Last name starts with a vowel.

   ```{r}
   str_subset(x, " [AEIOUaeiou]")
   ```

   ```{python}
   [s for s in x if re.search(r" [AEIOUaeiou]", s)]
   ```

3. Either the first or last name starts with a vowel.

   ```{r}
   str_subset(x, "^[AEIOU]| [AEIOUaeiou]")
   ```

   ```{python}
   [s for s in x if re.search(r"^[AEIOU]| [AEIOUaeiou]", s)]
   ```

4. Neither the first nor last name starts with a vowel.

   ```{r}
   str_subset(x, "^[AEIOU]| [AEIOUaeiou]", negate = TRUE)
   ```

   ```{python}
   [s for s in x if not re.search(r"^[AEIOU]| [AEIOUaeiou]", s)]
   ```


# Exercise 2

Write a regular expression that extracts all of the phone numbers in `text`, then use groups to extract the area code separately from the rest of the number.

```{r}
text = c(
  "apple",
  "219 733 8965",
  "329-293-8753",
  "Work: (579) 499-7527; Home: (543) 355 3679"
)
```

```{python}
text = r.text
```

1. Extract all phone numbers.

   ```{r}
   str_extract_all(text, "\\(?\\d{3}\\)?[- ]\\d{3}[- ]\\d{4}")
   ```

   ```{python}
   [re.findall(r"\(?\d{3}\)?[- ]\d{3}[- ]\d{4}", s) for s in text]
   ```

2. Area code separately from the rest of the number.

   ```{r}
   str_match_all(text, "\\(?(\\d{3})\\)?[- ](\\d{3}[- ]\\d{4})")
   ```

   ```{python}
   [re.findall(r"\(?(\d{3})\)?[- ](\d{3}[- ]\d{4})", s) for s in text]
   ```
